% ============================================================
% abstract.tex
% ============================================================

We propose that conservation of probability --- the single axiom $\mathcal{A}=1$
--- is sufficient to derive the principal results of quantum mechanics without
additional postulates.
On a bipartite three-dimensional causal lattice $\Tdiamond$,
each particle is represented as a causal session: an independent phase oscillator
that propagates amplitude across the lattice subject to $|\psi_R|^2 +
|\psi_L|^2 = 1$ at every tick.
No Hamiltonian, no Hilbert space, and no measurement postulate are assumed.

From this single constraint we derive:
\textbf{(i)} the Dirac equation in the continuum limit $a\to 0$, with the
gamma matrices identified as geometric encodings of the RGB/CMY bipartite
sublattice geometry; rotational invariance is recovered on
$O_h$-averaged observables (the same mechanism by which staggered fermions
recover Lorentz invariance in lattice QCD), and the residual lattice
birefringence aligned with the optical axis
$\mathbf{V}_1+\mathbf{V}_2+\mathbf{V}_3 = (1,1,-1)$ is a falsifiable
prediction with closed-form coefficients (Section~\ref{sec:kinematics});
\textbf{(ii)} wave-particle duality and the Born rule as consequences of
amplitude redistribution under $\mathcal{A}=1$;
\textbf{(iii)} quantization of atomic orbitals as Arnold tongue attractors of
the two-body phase dynamics, recovering the hydrogen Bohr radius to four
significant figures without fitting;
\textbf{(iv)} gravity as a clock-density gradient rather than spacetime
curvature --- a refraction of causal paths through regions of higher
session density, reproducing the $1/r^2$ force law with a falsifiable
discrete correction at short range;
\textbf{(v)} photon emission as a structural consequence of $\mathcal{A}=1$:
the Coulomb interaction is a probability source that a single session cannot
accommodate, mandating creation of a new photon session at orbit lock-in;
partial numerical verification (\texttt{exp\_19c}), with a full multi-rate
scan pending.

Energy conservation and momentum conservation are not independent inputs ---
they are consequences of $\mathcal{A}=1$ applied to closed session sets.
The gauge structure of the Standard Model is reframed as the minimal set of
session types required to maintain $\mathcal{A}=1$ on the bipartite lattice
geometry.

The numerically tractable results are verified by a suite of more
than fifteen experiments with no free parameters beyond the
Planck-scale calibration of the lattice spacing; the framework's
remaining derivations --- black-hole thermodynamics among them --- are
established analytically under explicitly stated modelling
assumptions and flagged as such where they appear.

The particle content of the Standard Model is the catalogue of probability
imbalances the bipartite lattice geometry generates and the session types
$\mathcal{A}=1$ requires to resolve them.
The Standard Model is not a theory --- it is an accounting system.
$\mathcal{A}=1$ is the account.
