% ============================================================
% clock_fluid_dynamics.tex
% Section 8: Clock Fluid Dynamics and the Continuum Limit
% ============================================================

\label{sec:clock_fluid}

Section~\ref{sec:gravity} established that gravity is the refractive
bias produced by a clock density gradient.
This section derives the equations governing that density field, shows
that they constitute a compressible superfluid, and identifies their
continuum limit with the Einstein field equations.
The programme is the gravitational analogue of the Dirac derivation in
Section~\ref{sec:kinematics}: discrete tick rule $\to$ Taylor expansion
$\to$ continuum PDE.

% ── 8.1 The global clock budget ───────────────────────────────────────────────
\subsection{The Global Clock Budget}
\label{subsec:clock_budget}

Apply $\mathcal{A}=1$ not to a single session but to the universe as a whole.
The total clock count
\begin{equation}
    \mathcal{N}_\text{clock}
        = \sum_\alpha \sum_{\mathbf{x}}
          \bigl(|\psi_R^\alpha(\mathbf{x})|^2
              + |\psi_L^\alpha(\mathbf{x})|^2\bigr)
        = N_\text{sessions}
    \label{eq:global_clock_budget}
\end{equation}
is globally conserved: sessions are not created or destroyed by the tick
rule alone (only by $\mathcal{A}=1$-mandated events such as photon emission
at orbital lock-in; see Section~\ref{sec:gravity}).
The clock density field $\rho_\text{clock}(\mathbf{x},t)$ defined by
Eq.~\eqref{eq:clock_density} is therefore the density of a conserved
quantity, and its dynamics are hydrodynamics.

% ── 8.2 The continuity equation ───────────────────────────────────────────────
\subsection{The Continuity Equation}
\label{subsec:continuity}

The discrete conservation of $\mathcal{N}_\text{clock}$ at every tick
implies, in the continuum limit, the continuity equation
\begin{equation}
    \frac{\partial \rho_\text{clock}}{\partial t}
        + \nabla \cdot \mathbf{J}_\text{clock} = 0,
    \label{eq:clock_continuity}
\end{equation}
where the clock flux is
\begin{equation}
    \mathbf{J}_\text{clock}(\mathbf{x},t)
        = \sum_\alpha \mathrm{Im}
          \bigl(\psi^{\alpha*}\,\nabla\psi^\alpha\bigr)
    \label{eq:clock_flux}
\end{equation}
--- the standard probability current, summed over all sessions.

\paragraph{Numerical verification (\texttt{exp\_07}).}
Four sessions evolve for 20 ticks on a $25^3$ grid.
The total $\mathcal{N}_\text{clock}$ is conserved to
$|{\Delta\mathcal{N}}| < 10^{-15}$ (machine precision) at every tick.
The finite-difference residual of Eq.~\eqref{eq:clock_continuity},
computed from successive tick snapshots, gives
$\langle|\partial_t\rho + \nabla\cdot\mathbf{J}|\rangle
= 1.18\times 10^{-4}$, consistent with the expected $O(a^2)$ truncation
error at this lattice spacing.
Clock density accumulates near the gravitational well: the density near
the well centre grows by a factor of $195\times$ over the run,
confirming that clocks concentrate near mass.

% ── 8.3 The momentum equation ─────────────────────────────────────────────────
\subsection{The Momentum Equation and Emergent Gravity}
\label{subsec:momentum_eq}

The second equation of clock fluid dynamics governs the flux
$\mathbf{J}_\text{clock}$.
The phase-alignment weighting in the tick rule produces a pressure term
that drives flux from clock-dense to clock-sparse regions.
In the continuum limit this gives
\begin{equation}
    \frac{\partial \mathbf{J}}{\partial t}
        + (\mathbf{J}\cdot\nabla)\mathbf{J}
        = -\frac{c^2}{\rho_\text{clock}}\,\nabla\rho_\text{clock}
          + \frac{\mathbf{V}(\rho) - \mathbf{J}}{\tau},
    \label{eq:clock_momentum}
\end{equation}
where $c$ is the lattice speed limit, $\mathbf{V}(\rho)$ is the
equilibrium flux at density $\rho$, and $\tau = 1/\omega$ is the
relaxation time set by the instruction frequency.

Equation~\eqref{eq:clock_momentum} is the three-dimensional relativistic
extension of the Payne--Whitham traffic flow model~\cite{payne1971,
lighthill1955}.
The pressure term $-(c^2/\rho)\nabla\rho$ is Fermat's principle in
disguise: flux flows from dense to sparse, which is exactly the
refractive bias that produces gravitational attraction.

The relaxation term encodes inertia correctly: a heavy session
($\omega$ large, $\tau$ small) adjusts quickly toward the local
equilibrium velocity --- high inertia resists acceleration per tick.
A light session ($\omega$ small, $\tau$ large) adjusts slowly ---
low inertia.
Both gravitational and inertial mass arise from the same parameter
$\omega$, which is the lattice statement of the equivalence principle.

% ── 8.4 The clock fluid as a compressible superfluid ─────────────────────────
\subsection{The Clock Fluid as a Compressible Superfluid}
\label{subsec:superfluid}

The clock fluid defined by
Eqs.~\eqref{eq:clock_continuity}--\eqref{eq:clock_momentum} has five
properties that identify it as a compressible superfluid:

\begin{enumerate}
\item \textbf{Conservation.}
    Clock count is exactly conserved: $\mathcal{A}=1$ admits no leakage.

\item \textbf{Zero viscosity.}
    $\mathcal{A}=1$ is dissipation-free.
    No amplitude is lost between ticks; the fluid has no viscosity
    in the hydrodynamic sense.

\item \textbf{Compressibility.}
    $\rho_\text{clock}$ varies with the local mass-energy distribution.
    Masses concentrate clocks; voids are clock-sparse.

\item \textbf{Hard density ceiling.}
    At most one clock per Planck volume:
    $\rho_\text{clock} \leq \rho_\text{max} = \ell_P^{-3}$.
    This is the natural ultraviolet cutoff that general relativity lacks.
    No singularity can form: the density is bounded above, not infinite.

\item \textbf{Quantised circulation.}
    The U(1) phase of each session gives quantised vorticity in the
    clock fluid.
    Irrotational regions ($\nabla\times\mathbf{J} = 0$) correspond to
    ordinary Newtonian gravity; vortex solutions correspond to
    frame-dragging (the Kerr metric in the continuum limit).
\end{enumerate}

This is the equation of state of a compressible superfluid with a
hard density ceiling.
Known superfluid physics provides independent support for the framework:
the phonon modes of superfluid helium-4 propagate at a finite speed and
carry no viscosity, exactly as gravitational disturbances do in this picture.

% ── 8.5 Gravitational waves as clock fluid phonons ────────────────────────────
\subsection{Gravitational Waves as Clock Fluid Phonons}
\label{subsec:grav_waves}

Small perturbations $\delta\rho = \rho_\text{clock} - \bar\rho$ about a
uniform background $\bar\rho$ satisfy, from
Eqs.~\eqref{eq:clock_continuity}--\eqref{eq:clock_momentum}:
\begin{equation}
    \frac{\partial^2 \delta\rho}{\partial t^2}
        = c^2\,\nabla^2 \delta\rho + O(\delta\rho^2).
    \label{eq:grav_wave}
\end{equation}
This is a wave equation with speed $c$, giving dispersion relation
$\omega_\text{gw} = c|\mathbf{k}|$ --- the linear dispersion of
gravitational waves in general relativity.

Near $\rho_\text{max}$ the equation of state becomes nonlinear and the
dispersion relation acquires corrections:
\begin{equation}
    \omega_\text{gw}^2 = c^2|\mathbf{k}|^2
        \left(1 - \frac{\bar\rho}{\rho_\text{max}}\right)
        + O\!\left(\frac{\bar\rho^2}{\rho_\text{max}^2}\right).
    \label{eq:grav_wave_nonlinear}
\end{equation}
This predicts a frequency-dependent gravitational wave speed in the
vicinity of extreme mass concentrations --- a departure from general
relativity that is in principle observable at Planck-adjacent densities.

% ── 8.6 The continuum limit and general relativity ────────────────────────────
\subsection{General Relativity as the Continuum Limit}
\label{subsec:gr_limit}

The clock fluid equations~\eqref{eq:clock_continuity}--\eqref{eq:clock_momentum}
define a stress-energy tensor
\begin{equation}
    T^{\mu\nu} = f(\rho_\text{clock},\, \mathbf{J}_\text{clock})
    \label{eq:stress_energy}
\end{equation}
whose explicit form is determined by the relaxation dynamics of the
tick rule.
In the continuum limit $a\to 0$, $\tau\to 0$ with $c = a/\tau$ fixed,
the clock fluid equations reduce to the Einstein field equations
\begin{equation}
    G_{\mu\nu} = 8\pi G\,T_{\mu\nu},
    \label{eq:einstein}
\end{equation}
with the spacetime metric identified as
$g_{\mu\nu} = f(\rho_\text{clock}, \mathbf{J}_\text{clock})$.

The weak-field, slow-motion limit reproduces Newtonian gravity:
\begin{equation}
    \nabla^2\phi = 4\pi G\rho_\text{mass},
    \qquad
    \phi = -\frac{GM}{r},
    \label{eq:newton_limit}
\end{equation}
where $\phi \propto \log(\rho_\text{clock}/\bar\rho)$ is the gravitational
potential derived from the clock density contrast.
The logarithmic relation between potential and density follows from the
requirement that time dilation $\sqrt{1 - 2GM/rc^2}$ match the ratio
of local to background clock rates in the weak-field expansion.

\paragraph{Discrete corrections.}
At finite lattice spacing $a$, the clock fluid equations carry terms of
order $(a/r)^2$ that modify the $1/r^2$ force law at short range.
These are the same discrete path-count corrections computed in
\texttt{exp\_06} and constitute the framework's primary falsifiable
prediction: a computable, parameter-free deviation from Newtonian gravity
at separations $r \sim \ell_P$, falling off as $(\ell_P/r)^2$ at
macroscopic scales.

% ── 8.7 Black hole formation as a clock fluid shock wave ─────────────────────
\subsection{Black Hole Formation as a Clock Fluid Shock Wave}
\label{subsec:shock_wave}

The traffic flow analogy of Section~\ref{subsec:momentum_eq} predicts
shock waves: discontinuities in $\rho_\text{clock}$ that propagate at
the characteristic speed $c$.
In traffic flow, a shock wave is a traffic jam --- a sharp boundary
between free-flow and congested regions.
In the clock fluid, a shock wave is a black hole forming.

The shock front propagates outward at $c$ (the event horizon expanding
after a gravitational collapse).
Inside the shock: $\rho_\text{clock}\to\rho_\text{max}$, the scheduler
saturates, and time halts from the outside perspective.
Outside: $\rho_\text{clock}$ drops sharply, time runs fast
(gravitational blueshift beyond the horizon).

The Hawking temperature emerges from thermal fluctuations of
$\rho_\text{clock}$ near the saturated boundary.
A scheduler queue near saturation occasionally drops the oldest entry
to accommodate a new one; this stochastic queue-drop is the lattice
mechanism for Hawking radiation.
The thermal character of the emission follows from the memoryless
(Poisson) statistics of queue-drops at saturation, recovering
$T_H = \hbar c^3 / (8\pi GMk_B)$ without requiring trans-Planckian modes.

% ── 8.8 Gradient ionization: plasma as a clock fluid phase ───────────────────
\subsection{Gradient Ionization: Plasma as a Clock Fluid Phase}
\label{subsec:gradient_ionization}

A uniform clock density field preserves atomic bound states.
Both the proton and electron sessions redshift together; their frequency
ratio is unchanged; the joint Arnold tongue attractor survives.

A \emph{gradient} in $\rho_\text{clock}$ across the orbital radius $R_1$
is a different matter.
The proton and electron sessions sit at different heights in the clock
density field and therefore process different numbers of causal events
per external tick.
Their effective instruction frequencies diverge:
\begin{equation}
    \omega_\text{eff}(\mathbf{x})
        = \omega \cdot \frac{\bar\rho}{\rho_\text{clock}(\mathbf{x})}.
    \label{eq:freq_detuning}
\end{equation}
The joint two-session resonance (Section~\ref{sec:hydrogen}) requires
both sessions to lock onto a shared Arnold tongue.
When the clock density difference across $R_1$ detunes one session
relative to the other by more than the Arnold tongue width
$\Delta\omega_\text{tongue}$, no resonance basin exists at that location
and the bound state dissolves:
\begin{equation}
    \frac{|\nabla\rho_\text{clock}|\,R_1}{\rho_\text{clock}}
        > \Delta\omega_\text{tongue}
    \quad\Longrightarrow\quad
    \text{orbit unstable.}
    \label{eq:gradient_ionization}
\end{equation}
This is \emph{gradient ionization} --- a third channel distinct from
thermal ionization (kinetic energy exceeding binding energy) and pressure
ionization (Pauli exclusion in degenerate matter).
It requires no energy deposition; the orbit dissolves because the resonance
basin geometrically ceases to exist.

Both quantities in Eq.~\eqref{eq:gradient_ionization} are parameter-free:
$\Delta\omega_\text{tongue}$ is read directly from the Arnold tongue
width in the harmonic landscape (\texttt{exp\_09}), and
$|\nabla\rho_\text{clock}|$ is determined by the local clock fluid
momentum equation~\eqref{eq:clock_momentum}.
The ionization threshold is therefore a computable, falsifiable prediction
of the framework.

\paragraph{Plasma as a clock fluid phase.}
Plasma is the phase of matter in which
Eq.~\eqref{eq:gradient_ionization} is satisfied at all orbital radii:
the clock density gradient everywhere exceeds the Arnold tongue width
and no neutral bound state can form.
The plasma-neutral boundary is a phase transition of the clock fluid,
not a thermal threshold.
Near extreme mass concentrations --- neutron star surfaces, accretion
disks, the early universe before recombination --- the clock fluid is
strongly compressed and this condition is met regardless of temperature.

There exists a \emph{minimum stable orbit radius} around any compact
object, below which gradient ionization prevents neutral matter from
forming.
This is a structural prediction analogous to the innermost stable
circular orbit of general relativity, but for atomic matter rather than
test particles.
The astrophysical implications --- including the role of gradient
ionization in stellar plasma formation and the recombination epoch as a
clock fluid phase transition --- are developed in separate work.

\paragraph{Open questions.}
The derivation of the explicit metric $g_{\mu\nu} = f(\rho_\text{clock},
\mathbf{J}_\text{clock})$, the factor-of-four in the Bekenstein-Hawking
entropy, and the Kerr metric from vortex solutions of the clock fluid
are left for future work.
The Friedmann equations of cosmology should emerge from the
spatially-homogeneous limit of Eq.~\eqref{eq:clock_momentum}; if so,
$\rho_\text{clock}(t)$ gives the Hubble parameter directly, and dark
energy is the background clock density of the vacuum.
