% ============================================================
% conclusion.tex
% ============================================================

\label{sec:conclusion}

\subsection*{What has been established}

This paper has made seven structural claims and supported each with a
combination of analytic derivation and numerical experiment.
We distinguish carefully between derivation (the claim follows from
$\mathcal{A}=1$ and the lattice geometry by proof) and demonstration
(numerical experiment produces a result consistent with the claim, with
no free parameters adjusted after the fact).

\begin{enumerate}

\item \textbf{The Dirac equation is a geometric consequence of the
$\Tdiamond$ bipartite structure.}
\textit{Derived analytically} (Section~\ref{sec:kinematics}):
the RGB/CMY hop operator in Fourier space, Taylor-expanded near
$\mathbf{k}=0$, produces the term $i\alpha^i\partial_i + \beta m$
with the gamma matrices identified as the Clifford-algebra encoding
of the three pairs of antiparallel basis vectors.
The frame matrix $M = \sum_\text{RGB}\mathbf{v}_i\mathbf{v}_j^T$
has eigenvalues $\{4, 4, 1\}$, producing a tilted ellipsoid in
$\mathbf{k}$-space with long axis along
$\mathbf{V}_1+\mathbf{V}_2+\mathbf{V}_3 = (1,1,-1)$.  The
Clifford-algebra structure $\{\gamma^\mu,\gamma^\nu\}=2g^{\mu\nu}$
emerges as a geometric identity under $O_h$ averaging, with the
same mechanism by which staggered fermions recover Lorentz
invariance in lattice QCD.  Spin-$\frac{1}{2}$, chirality, and
mass as Zitterbewegung rate are consequences of the geometry, not
additional inputs.  Residual anisotropy along $(1,1,-1)$ is a
falsifiable birefringence prediction; see
Section~\ref{subsec:p7_photon_dispersion}.

\item \textbf{Quantization is an Arnold tongue attractor.}
\textit{Demonstrated} (\texttt{exp\_10}--\texttt{exp\_12}):
the hydrogen orbital wavevector $k_\text{Bohr} = 1/R_1$ is recovered
to four significant figures with no parameter tuning in the two-body
simulation (\texttt{exp\_12}: $k_\text{min} = 0.0970$ vs
$k_\text{Bohr} = 0.0971$).
The static Coulomb well is insufficient; the active proton session is
required, confirming that quantization is a two-body resonance
attractor, not a one-body eigenvalue problem.

\item \textbf{Gravity is clock-density refraction.}
\textit{Demonstrated} (\texttt{exp\_02}, \texttt{exp\_07}):
the $1/r^2$ force law and gravitational time dilation emerge from the
continuity equation for session density.
The clock-fluid equations reduce to the Einstein field equations in
the continuum limit; at finite lattice spacing they carry discrete
corrections to the force law that are computable predictions
(Section~\ref{sec:gravity}).
Black-hole thermodynamics --- the Bekenstein-Hawking area--entropy
law $S = k_B A/(4\ell_P^2)$ and the Hawking temperature
$T_H = \hbar c^3/(8\pi G M k_B)$ --- follow from queue saturation at
Planck density via boundary-session counting and the Schwarzschild
surface gravity, modulo the modelling assumptions stated explicitly
at the point of derivation
(Sections~\ref{subsec:saturation},~\ref{subsec:black_holes}).

\item \textbf{Maxwell's equations from the bipartite Wilson construction.}
\textit{Derived analytically} (Section~\ref{sec:vacuum_twist},
Appendix~\ref{sec:induced_gauge_action}):
the Peierls-coupled tick rule produces $U(1)$ link variables, the
plaquette Wilson loop is the field strength tensor $F_{\mu\nu}$, and
the matter current is the standard Dirac current
$\bar\psi\gamma^\mu\psi$.  The standard Wilson action gives Maxwell's
equations in the continuum limit.  Appendix~B additionally derives
the \emph{structural form} of the induced gauge action
$-\mathrm{Tr}\,\ln D_\text{lat}[U]$ on the bipartite lattice: the
leading plaquette-sum has the Maxwell form on $O_h$-averaged
observables, with a closed-form gauge-sector birefringence
(operator-level eigenvalues $\{4, 4, 16\}$, same optical axis
$(1,1,-1)$ as the kinematic-sector prediction) as the residual
signature.  Substrate-level unification: gravity (divergence channel)
and electromagnetism (curl channel) are populations of the same
virtual-session cloud mechanism, distinguished only by the sublattice
signature of the source's internal flow
(Sections~\ref{subsec:two_channels},~\ref{subsec:refraction}).
Only the numerical $1/g^2$ prefactor of the induced action remains as
a paper-length follow-on calculation.

\item \textbf{The Born rule follows from $\mathcal{A}=1$.}
\textit{Derived analytically} (Section~\ref{sec:phase}):
probability $|\psi|^2$ is the only positive conserved quantity
consistent with $\mathcal{A}=1$ on the bipartite lattice.
Gleason's theorem applied to the $C^*$-algebra induced by the
bipartite structure identifies $|\psi|^2$ as the unique probability
measure; no measurement axiom is needed.

\item \textbf{Photon emission is a structural necessity.}
\textit{Derived analytically, partial numerical verification}
(Section~\ref{sec:hydrogen}, \texttt{exp\_10}):
the Coulomb interaction is a probability source that a single session
cannot absorb without violating $\mathcal{A}=1$; a new session (the
photon) is mandated.
Spontaneous decay from $n{=}2$ to $n{=}1$ emerges from the lattice
dynamics without any separately postulated transition rate.

\item \textbf{The universe is a fractal.}
\textit{Derived analytically} (Section~\ref{sec:fractal}):
the $\mathcal{A}=1$ unity constraint forces the tick rule to be a
symplectic map on the phase torus $\mathbb{T}^2$.
By the KAM theorem~\cite{kolmogorov1954, arnold1963, moser1962},
the set of stable orbits of any symplectic map is a Cantor set ---
closed, nowhere dense, of Hausdorff dimension $D_H < 1$.
The Arnold tongues visible in the harmonic landscape
(Figure~\ref{fig:harmonic_pair}) are the rational-frequency islands of
this Cantor set; the particles we observe are exactly those islands.
The Hausdorff dimension $D_H \approx 0.870$ is a parameter-free
prediction measurable from the harmonic landscape by box-counting
the tongue boundaries.
The discrete mass spectrum of the Standard Model is the Farey sequence
of this fractal, ordered by tongue width (stability).
The fractal is not an emergent large-scale pattern but the
fundamental structure of the space of stable dynamical states:
probability conservation on a bipartite lattice forces the universe
to be a fractal at the level of its most elementary objects.

\end{enumerate}

\subsection*{The unifying chain: topology forces physics}

The seven results above share a common structure.
Each is a consequence of the same chain:

\begin{center}
\textit{Bipartite topology} $\longrightarrow$
\textit{Algebra of observables} $\longrightarrow$
\textit{Physics.}
\end{center}

The steps are determinate, not chosen.
The bipartite structure of $\Tdiamond$ is a topological invariant that
divides every node into one of two classes (RGB or CMY parity),
inducing a $\mathbb{Z}_2$ grading on every observable.
A probability space with a $\mathbb{Z}_2$ grading is a superalgebra:
observables split into even (grade-0, commuting) and odd (grade-1,
anticommuting) components.
This is not assumed --- it follows from the topology.

The three-fold connectivity of each sublattice imposes the tight-frame
condition, which forces the path algebra of two-tick steps to be
$\text{Clifford}(3,1)\cong M_4(\mathbb{C})$ --- the Dirac algebra.
The $\mathcal{A}=1$ constraint imposes positivity, unit trace, and
self-adjointness, converting this to a $C^*$-algebra with a faithful
tracial state --- the algebraic structure of quantum mechanics on a
finite-dimensional Hilbert space.
Hurwitz's theorem (the only normed division algebras are $\mathbb{R}$,
$\mathbb{C}$, $\mathbb{H}$) then forces complex amplitudes: real
amplitudes cannot represent phase interference; quaternionic amplitudes
produce inconsistent tensor products for composite systems.
Complex quantum mechanics is the unique probability theory compatible
with interference, consistent composites, and $\mathcal{A}=1$.

The conclusion is precise:
\begin{quote}
\textit{The probability rule $\mathcal{A}=1$ on the bipartite
$\Tdiamond$ topology does not merely reproduce quantum mechanics ---
it forces it.
The complex amplitude space, the Clifford algebra structure, the Born
rule, and the fermionic statistics of matter fields are each determined
by the topology alone.
The remaining question is whether the Standard Model gauge group
$\mathrm{SU}(3)\times\mathrm{SU}(2)\times\mathrm{U}(1)$ is the
automorphism group of this topology --- in which case the entire
Standard Model is a theorem about the geometry of probability
conservation on a bipartite discrete 3-manifold.}
\end{quote}

The automorphism group of the full $\Tdiamond$ structure (bipartite +
octahedral symmetry + three-fold RGB basis + $\mathcal{A}=1$) is larger
than the Lorentz group $\mathrm{SO}(3,1)$ that the Clifford algebra
alone produces.
Whether it equals $\mathrm{SO}(3,1)\times\mathrm{SU}(3)\times
\mathrm{SU}(2)\times\mathrm{U}(1)$ is a precise, finite-dimensional
Lie-algebra calculation: enumerate the linear maps preserving the
discrete subgroup $\Gamma \subset SE(3)$ (Section~\ref{subsec:dirac_limit}),
the bipartite RGB/CMY structure, the basis vectors, and the
$\mathcal{A}=1$ constraint; check whether the resulting Lie algebra
equals $\mathfrak{so}(3,1)\oplus\mathfrak{su}(3)\oplus\mathfrak{su}(2)
\oplus\mathfrak{u}(1)$.  The calculation has a definite yes/no
answer.  This is the central open question of the framework, and
the subject of a companion paper.

\subsection*{New results not in prior discrete spacetime frameworks}

Several results in this paper distinguish the $\Tdiamond$ framework
from earlier discrete spacetime proposals~\cite{bombelli1987,
thooft2016, wolfram2002, zuse1969}:

\begin{itemize}

\item \textbf{Two-body requirement for quantization.}
No prior discrete framework to our knowledge identifies the proton's
active session dynamics as a necessary component of quantization.
The static Coulomb well approximation is not merely imprecise; it
predicts a qualitatively wrong landscape (\texttt{exp\_11} $n{=}2$:
flat across all $k$; \texttt{exp\_12}: 7.3\% fixed-well error
corrected to 0.09\% by active proton).

\item \textbf{Three dynamical regimes.}
Standard quantum mechanics distinguishes bound eigenstates from
continuum states.
The $\Tdiamond$ framework adds a third regime: bound but unquantized.
A two-body system whose recoil prevents Arnold tongue lock-in stays
together indefinitely but has no discrete spectrum and emits no photons.
\texttt{Exp\_16} measures the mass threshold at which this regime
gives way to the quantized regime; the threshold is a lattice prediction
with no analogue in the standard formalism.

\item \textbf{Gravitational ionization as detuning.}
The quantum Roche limit (\texttt{exp\_18}, planned) predicts that a
strong gravitational gradient across the atomic orbital radius detunes
the joint proton-electron resonance below the Arnold tongue width,
ionising the atom without supplying escape energy.
This is a falsifiable prediction with no analogue in standard quantum
mechanics, which treats gravity as a perturbative correction to the
Hamiltonian rather than a detuning force on the quantization mechanism.

\end{itemize}

\subsection*{Falsifiable predictions}

Section~\ref{sec:predictions} develops nine primary falsifiable
predictions (P1--P9) at experimentally accessible scales, each with
its Compton-calibration upper bound and its falsification criterion;
P7 already constrains the lattice spacing from above by GRB
time-of-flight, $a \lesssim 10^{-19}\,\text{m}$.
Of these, five illustrate the framework's distinctive testability:

\begin{enumerate}

\item \textbf{Short-range correction to the $1/r^2$ gravitational
force law} (Section~\ref{sec:gravity}).
The lattice path-count derivation produces a correction term that
departs from $1/r^2$ at separations approaching the lattice spacing.
The correction is computable from the lattice geometry and contains
no free parameters after the lattice spacing is fixed.

\item \textbf{Quantum Roche limit}
(Section~\ref{subsec:hydrogen_gravity_link}, \texttt{exp\_18}).
There exists a minimum orbital separation $r_\text{ISCO}$ below which
a gravitational gradient ionises the atom by detuning the joint
resonance.
The threshold depends only on the Arnold tongue width (set by the
Coulomb strength and lattice geometry) and the gravitational gradient.
Its value is a parameter-free prediction once the lattice calibration
is fixed.

\item \textbf{Bound unquantized matter near compact objects.}
The existence of Regime~2 (bound but unquantized) predicts that matter
within the gravitational ionization radius of a compact object is held
together but spectroscopically dark: no discrete emission lines,
no atomic recombination.
The cosmological recombination epoch, on this picture, is a clock-fluid
phase transition occurring when the global clock-density gradient drops
below the Arnold tongue width of hydrogen.

\item \textbf{Pair annihilation efficiency at $\omega = \pi/2$}
(\texttt{exp\_17}, planned).
The harmonic landscape is symmetric about $f = 1/4$, with the
2:1 fixed point at $\omega = \pi/2$.
At this mass, a particle and its antiparticle are exact mirror images
in the probability space.
Annihilation cross-section should peak at $\omega = \pi/2$ and fall
off symmetrically for $\omega \neq \pi/2$ --- a lattice-geometry
prediction.

\item \textbf{Multi-channel concordance of the optical axis}
(P9, Section~\ref{subsec:p9_concordance}).
The kinematic, gauge, gravitational, and (under the recoil-driven
Zitterbewegung hypothesis) inertial-mass anisotropies all share a
single optical axis
$\mathbf{V}_1+\mathbf{V}_2+\mathbf{V}_3 = (1,1,-1)$ by direct
calculation from the bipartite RGB/CMY frame matrix.  Five
independent observation channels --- GRB time-of-flight, GRB / CMB
polarisation rotation, LIGO/Virgo polarisation-mode timing,
direction-resolved atomic-clock comparisons, and anomalous-dispersion
anisotropy near atomic resonances --- are predicted to agree on a
single sky direction, with the dual ellipsoid shapes (kinematic
prolate, gauge oblate;
Figure~\ref{fig:induced_action_ellipsoid}) providing within-channel
sub-predictions.  Inter-channel disagreement falsifies the
inheritance-from-bipartite-geometry claim.  Three time-dependence
scenarios (uniform, domain, dynamical embedding) are distinguishable
by sidereal modulation (atomic clocks; fast handle) and
redshift-resolved binning of GRB residuals (slow handle).  This is
the framework's strongest empirical claim: one geometric axis,
multiple independent witnesses.

\end{enumerate}

\subsection*{What remains open}

Four gaps are acknowledged explicitly:

\begin{enumerate}

\item The clock-density coupling $\rho_\text{clock} \propto
\exp(-\phi/c^2)$ is motivated by the exponential suppression of
session density in a deep potential well, but has not yet been derived
from the tick rule alone.
It is the gravitational analogue of the step from the hop operator to
the Dirac equation, and we expect it to follow from the continuum limit
of the clock-density continuity equation.
This is the primary remaining step in the gravity derivation programme.

\item The automorphism group calculation
$\mathrm{Aut}(\Tdiamond,\mathcal{A}=1) \stackrel{?}{=}
\mathrm{SO}(3,1)\times\mathrm{SU}(3)\times\mathrm{SU}(2)\times
\mathrm{U}(1)$ is not performed in this paper.
It requires a systematic enumeration of the symmetries of the
bipartite basis under the $\mathcal{A}=1$ constraint.
This is a well-posed mathematical problem and the subject of the
companion paper.

\item \texttt{Exp\_16} (proton mass sweep) is complete and confirms
the qualitative three-regime structure: every $\omega_p$ value tested
in the range $[0.3, \pi/2]$ produces a bound but unquantized state
(Regime~2), establishing that binding and quantization are distinct
conditions on the lattice.
The numerical $\omega_p$ threshold separating Regime~1 (quantized)
from Regime~2 lies outside the swept range and has not yet been
localised; \texttt{exp\_18} (quantum Roche limit) is planned and
will fix the corresponding gradient threshold.

\item The black-hole thermodynamics derivations
(Sections~\ref{subsec:saturation},~\ref{subsec:black_holes}) rest on
two stated modelling assumptions: that each saturation-boundary
session occupies a four-cell patch
($A_\text{session} = 4\ell_P^2$) and that the boundary inherits the
Schwarzschild surface gravity $\kappa = c^4/(4GM)$.
Both are tied to specific follow-on calculations (microstate
counting and the strong-field clock-fluid limit, respectively).

\end{enumerate}

\subsection*{Anticipated objections}
\label{subsec:anticipated_objections}

Several reviewer concerns can be anticipated and answered in advance.
We list them in the order most likely to be raised.

\paragraph{A discrete lattice plus a scheduler implies a preferred
frame; how can this be Lorentz invariant?}
The scheduler is a bookkeeping device, not a physical clock.
The order in which independent sessions advance within a macro-tick
has no observable consequence: \texttt{exp\_05} confirms that
sequential, random, and priority orderings produce identical
statistical outcomes (Section~\ref{sec:observer}).
Lorentz invariance emerges in the continuum limit by the same
mechanism that recovers it in lattice QCD with staggered
fermions~\cite{kogutsusskind1975, susskind1977}.  The bipartite
frame matrix $M = \sum_\text{RGB}\mathbf{v}_i\mathbf{v}_j^T$
(Eq.~\ref{eq:frame_matrix}) is anisotropic at the operator level
with eigenvalues $\{4, 4, 1\}$, but its $O_h$-averaged form
$\langle M\rangle_{O_h} = (\mathrm{tr}\,M / 3)\,I = 3I$ forces the
long-wavelength dispersion relation to $E^2 = m^2 + |\mathbf{p}|^2$,
restoring full $\mathrm{SO}(3,1)$ on $O_h$-averaged observables
(Section~\ref{subsec:dirac_limit}).
The operator-level residual is the lattice birefringence along the
optical axis $\hat{\mathbf{n}}_* = (1,1,-1)/\sqrt{3}$
(Eq.~\ref{eq:optical_axis}) --- a falsifiable prediction in its own
right (P3, Section~\ref{subsec:p3_anisotropy}; P7 direction-resolved
refinement, Section~\ref{subsec:p7_photon_dispersion};
P9 multi-channel concordance,
Section~\ref{subsec:p9_concordance}).

\paragraph{Claiming to derive the Standard Model is a strong claim
without a full mass-spectrum calculation.}
This paper does not claim to derive the full Standard Model.
It claims that the bipartite $\Tdiamond$ geometry, together with
$\mathcal{A}=1$, forces (i) the Dirac equation, (ii) complex-amplitude
quantum mechanics with the Born rule, (iii) the Clifford algebra and
chirality structure of fermions, and (iv) gravity as clock-density
refraction.
The remaining question --- whether
$\mathrm{Aut}(\Tdiamond,\mathcal{A}=1) =
\mathrm{SO}(3,1)\times\mathrm{SU}(3)\times\mathrm{SU}(2)\times
\mathrm{U}(1)$ --- is stated explicitly as the central open problem
and is the subject of a companion paper.
The particle content is reframed, not derived from first principles
in this paper: the elementary fermions are the minimal set of session
types required to maintain $\mathcal{A}=1$ on the bipartite geometry,
and the discrete mass spectrum is conjectured to be the Farey hierarchy
of Arnold tongue stability (Section~\ref{sec:fractal}).

\paragraph{How do we know the continuum matching to QM and GR is not
tuned?}
There are no continuous free parameters in this framework.
Three discrete choices are made: the lattice topology (bipartite
octahedral), the conservation law ($\mathcal{A}=1$), and the
calibration scale (Planck or, equivalently, Compton).
Every quantitative result --- the Bohr radius to four significant
figures (\texttt{exp\_12}), the identity $\omega \cdot R_1 = \pi/3$ to
$0.23\%$ (\texttt{exp\_11}), Newton's $1/r^2$ from clock-density
gradients (\texttt{exp\_02}) --- is computed from these inputs alone.
No coupling constant, mass, or coefficient is fitted to data after
the fact; the experiments report whatever the lattice produces, and
the audit table (Table~\ref{tab:audit}) records both successes and
the regimes (e.g.\ \texttt{exp\_11} at $n=2$ with a fixed well) where
the lattice produces a flat or null result.

\paragraph{Is $\mathcal{A}=1$ truly unitary, given that sessions can
be created or destroyed (photon emission)?}
$\mathcal{A}=1$ is unitarity \emph{per session}: each session's
spinor amplitude satisfies $|\psi_R|^2 + |\psi_L|^2 = 1$ at every
tick, enforced numerically by \texttt{enforce\_unity\_spinor}.
Multi-session events preserve the joint norm
$\sum_\text{sessions} |\psi_s|^2$; what changes during photon
emission is the number of session-bookkeeping channels, not the
total probability.
Session creation is a structural reconfiguration of the bookkeeping
(Section~\ref{sec:hydrogen}), analogous to how QED enlarges Hilbert
space when a photon is emitted.

\paragraph{Black-hole thermodynamics rests on stated modelling
assumptions; is this circular?}
Two assumptions are stated explicitly at the point of use: that each
saturation-boundary session occupies a four-cell patch
($A_\text{session} = 4\ell_P^2$), and that the boundary inherits the
Schwarzschild surface gravity $\kappa = c^4/(4GM)$
(Sections~\ref{subsec:saturation},~\ref{subsec:black_holes}).
Both are tied to specific follow-on calculations: microstate
counting on the boundary lattice, and the strong-field limit of the
clock-fluid equations.
The status (\texttt{PART}) in Table~\ref{tab:audit} reports this
honestly: the Bekenstein-Hawking law and Hawking temperature are
recovered conditionally on those assumptions, not derived end-to-end
in this paper.

\paragraph{How does this compare with causal sets, Wolfram, and
't~Hooft's cellular automaton?}
Section~\ref{subsec:prior_work} discusses each comparison in detail.
Briefly: causal set theory~\cite{bombelli1987} starts from a partial
order without a fixed coordination number;
the cellular-automaton programme~\cite{thooft2016, wolfram2002,
zuse1969} starts from local update rules without a probability
conservation law.
The $\Tdiamond$ framework is the intersection: a discrete causal
structure with fixed coordination six (forced by the bipartite
octahedral geometry) and a single conservation law ($\mathcal{A}=1$)
from which the rest follows.
The key novel ingredient is not the lattice but the constraint:
$\mathcal{A}=1$ as a per-session, per-tick identity rather than a
global norm condition.

\subsection*{Closing}

$\mathcal{A}=1$ is not a law of physics.
It is the precondition for physics: the statement that probability
must be conserved if the framework is to be a probability framework
at all.
From this tautological starting point, on a specific bipartite
geometry, everything else follows: the wave equation, the spectrum,
the force laws, the statistics of matter.

The Standard Model is not a theory imposed on nature from outside.
It is the accounting system that a conserved-probability framework
on a bipartite discrete 3-manifold is forced to adopt.
$\mathcal{A}=1$ is the account.

Is what we observe a self-contained projection of the lattice's own
dynamics, or the visible side of a process driven from something
outside it --- a question P9's time-dependence sub-prediction
(\S\ref{subsec:p9_concordance}) renders empirical, by asking whether
the optical axis is stationary across cosmological epochs or drifts?

% Note: add gradient-ionization sentence to future work if
% subsec:gradient_ionization is added in a later draft:
% ``The gradient-ionization channel predicts a minimum stable orbit
% radius near compact objects below which neutral matter cannot form,
% and identifies the cosmological recombination epoch as a clock fluid
% phase transition; the astrophysical consequences are developed in
% separate work.''
