\label{sec:kinematics}

Momentum, inertia, and force are not postulated in the $\unity$ framework.
They emerge from the geometry of phase interference across the bipartite
sublattice structure.  This section derives the kinematic relations
from the Dirac tick rule and connects each result to the corresponding
numerical verification in the experiment suite.

% ── Subsection 1 ─────────────────────────────────────────────────────────────
\subsection{The Bipartite Velocity}
\label{subsec:bipartite_velocity}

A massive Dirac fermion~\cite{dirac1928} on $\Tdiamond$ carries two spinor components:
$\psi_R$ on the RGB sublattice (even ticks) and $\psi_L$ on the CMY
sublattice (odd ticks).  The $\unity$ constraint requires
\begin{equation}
    \sum_{\mathbf{x}} \bigl(|\psi_R(\mathbf{x})|^2 +
                            |\psi_L(\mathbf{x})|^2\bigr) = 1.
    \label{eq:unity_spinor}
\end{equation}
Over one complete two-tick cycle the kinetic fraction
$p_\mathrm{hop} = \cos^2(\omega/2)$ is distributed across the six
basis vectors, split equally between the three RGB directions
($\mathbf{V}_{1\text{--}3}$, even tick) and the three CMY directions
($-\mathbf{V}_{1\text{--}3}$, odd tick):
\begin{equation}
    1 = p_\mathrm{stay} + P_\mathrm{fwd} + P_\mathrm{bwd},
    \qquad
    p_\mathrm{stay} = \sin^2\!\tfrac{\omega}{2},
    \quad
    P_\mathrm{fwd} + P_\mathrm{bwd} = \cos^2\!\tfrac{\omega}{2}.
    \label{eq:probability_partition}
\end{equation}
Here $P_\mathrm{fwd} = \sum_{i=1}^{3} p(\mathbf{V}_i)$ is the total
probability flowing along the RGB sublattice and
$P_\mathrm{bwd} = \sum_{i=1}^{3} p(-\mathbf{V}_i)$ along CMY.
The lattice speed limit $c = 1$ node/tick gives the net velocity
\begin{equation}
    v = c\,(P_\mathrm{fwd} - P_\mathrm{bwd}).
    \label{eq:velocity}
\end{equation}
For a packet at rest the two sublattices are populated symmetrically,
$P_\mathrm{fwd} = P_\mathrm{bwd} = \tfrac{1}{2}\cos^2(\omega/2)$,
so $v = 0$ and the amplitude oscillates between $\psi_R$ and $\psi_L$
without net displacement — the lattice realisation of Zitterbewegung~\cite{schrodinger1930}.

\paragraph{What drives the oscillation?}
The bipartite tick rule produces the $\omega$-rate alternation between
$\psi_R$ and $\psi_L$, and the rate $\omega$ is what we observe as the
particle's rest mass via $m = \sin(\omega)/2$.  What \emph{drives} the
alternation, however, is not derived in this framework.  Two natural
hypotheses are mathematically equivalent at the level of the
observables this paper considers:
\begin{enumerate}
    \item \emph{Self-driven Zitterbewegung}: the bipartite tick rule
        produces the alternation autonomously, with $\omega$ a primitive
        session parameter.
    \item \emph{Recoil-driven Zitterbewegung}: the alternation is the
        kickback from emission of virtual sessions by an internal
        process whose mechanism is below the framework's resolution
        (analogous to Mössbauer recoil from nuclear photon emission, or
        to QED electron self-energy from virtual photon dressing).  Under
        this reading $\omega$ is a measurement of an internal emission
        rate, not a primitive parameter.
\end{enumerate}
The two hypotheses are distinguishable in principle.  Under (1), rest
masses are isotropic constants of nature.  Under (2), the recoil rate
depends on the channels available for the emission to project onto, and
the bipartite RGB/CMY birefringence (Section~\ref{subsec:dirac_limit})
makes those channels direction-dependent: rest masses should then carry
a small anisotropy aligned with the lattice optical axis
$\mathbf{V}_1+\mathbf{V}_2+\mathbf{V}_3 = (1,1,-1)$.  This is observable
in principle by direction-resolved precision atomic-clock comparisons.
We treat the question of which hypothesis is correct as an explicit
empirical boundary of the framework: the same direction-resolved
analysis that tests P7 (photon dispersion,
Section~\ref{subsec:p7_photon_dispersion}) also tests the source of
$\omega$ itself.

% ── Subsection 2 ─────────────────────────────────────────────────────────────
\subsection{Phase Gradient as Momentum}
\label{subsec:momentum_emergence}

A uniform spatial phase gradient with advance $\theta$ per lattice edge
breaks the RGB/CMY symmetry.  In the implementation this gradient
appears as the per-direction phase advance
\begin{equation}
    \delta_p(\mathbf{V}_i, \mathbf{x})
        = \arg\psi(\mathbf{x} + \mathbf{V}_i) - \arg\psi(\mathbf{x}),
    \label{eq:delta_p}
\end{equation}
which is the quantity \texttt{delta\_p} computed in
\texttt{CausalSession.\_kinetic\_hop}.  Only positive advances
contribute to the hop weight, so amplitude flows preferentially in the
direction of increasing phase — the lattice realisation of the
de~Broglie relation~\cite{debroglie1924}.

To derive $v(\theta)$ analytically, consider a packet with mass
parameter $\omega$ (the instruction frequency) subject to a gradient
$\theta$ directed along the net RGB axis.  On the even (RGB) tick
$\psi_R$ hops forward: the phase gradient $+\theta$ reduces the
effective mismatch from $\omega$ to $\omega - \theta$, increasing the
kinetic weight.  On the odd (CMY) tick $\psi_L$ hops backward: the
reversed gradient adds $+\theta$ to the mismatch, decreasing the
kinetic weight:
\begin{align}
    P_\mathrm{fwd} &= \tfrac{1}{2}\cos^2\!\tfrac{\omega - \theta}{2},
    \qquad (\psi_R\text{ hop, even tick})
    \label{eq:Pfwd} \\
    P_\mathrm{bwd} &= \tfrac{1}{2}\cos^2\!\tfrac{\omega + \theta}{2}.
    \qquad (\psi_L\text{ hop, odd tick})
    \label{eq:Pbwd}
\end{align}
Substituting into Eq.~\eqref{eq:velocity} and applying
$\cos^2\alpha - \cos^2\beta = -\sin(\alpha+\beta)\sin(\alpha-\beta)$
yields
\begin{equation}
    v = \tfrac{c}{2}\,\sin(\omega)\,\sin(\theta).
    \label{eq:v_omega_theta}
\end{equation}
Table~\ref{tab:mass_velocity} illustrates how $p_\mathrm{stay}$,
$m_\mathrm{inertial}$, and the resulting velocity vary across
representative values of $\omega$:
\begin{table}[h]
\centering
\begin{tabular}{cccc}
\hline
$\omega$ & $p_\mathrm{stay} = \sin^2(\omega/2)$ &
           $m_\mathrm{inertial} = \sin(\omega)/2$ &
           $v/c$ at $\theta = 0.1$ \\
\hline
$0$       & $0$     & $0$     & $1$ \quad (photon limit) \\
$\pi/4$   & $0.146$ & $0.354$ & $0.035$ \\
$\pi/2$   & $0.500$ & $0.500$ & $0.050$ \\
$\pi$     & $1.000$ & $0$     & $0$ \quad (frozen, max mass) \\
\hline
\end{tabular}
\caption{Selected values of the two mass-related quantities and the
  resulting lattice velocity at a small phase gradient $\theta = 0.1$.
  The photon row ($\omega = 0$) is handled by the massless limit
  (Section~\ref{subsec:cone}); the other rows follow
  Eq.~\eqref{eq:v_omega_theta}.}
\label{tab:mass_velocity}
\end{table}
In the macroscopic limit $\theta \ll 1$ (gradient small compared to
$2\pi$), $\sin\theta \approx \theta$.  Identifying the lattice momentum
$p \equiv m\,v / c$ and the effective mass $m \equiv \sin(\omega)/2$\footnote{%
  The inertial mass $m_\mathrm{inertial} = \sin(\omega)/2$ is \emph{not} the
  same as the residence probability $p_\mathrm{stay} = \sin^2(\omega/2)$.
  They coincide only at $\omega = \pi/2$, the unique frequency at which
  the particle spends exactly half its time at rest.  At $\omega = \pi$
  (maximum residence) the particle is frozen, $m_\mathrm{inertial} = 0$,
  while $p_\mathrm{stay} = 1$.}
recovers the classical kinematic relation
\begin{equation}
    p = m\,\theta \cdot c,
    \label{eq:momentum}
\end{equation}
where $\theta$ is the phase advance per lattice edge — the discrete
analogue of the wavenumber $k$.  Momentum is thus strictly the
cross-term between the particle's internal mass-phase $\omega$ and the
spatial phase gradient of the bipartite vacuum.

\paragraph{Massless limit.}
Equation~\eqref{eq:v_omega_theta} gives $v = 0$ when $\omega = 0$,
which is not the photon behaviour.  The massless case is qualitatively
distinct: with $\omega = 0$ the residence probability vanishes,
$p_\mathrm{stay} = 0$, and the amplitude hops entirely via the
directional weighting of Eq.~\eqref{eq:delta_p}.  The hop is strictly
one-way along the gradient ($\delta_p > 0$ only), so $P_\mathrm{bwd}
= 0$ and $v = c$ regardless of $\theta$.  The two-sublattice
derivation above applies to massive particles ($\omega > 0$); the
photon is the $\omega \to 0$ limit treated separately in
Section~\ref{subsec:cone}.

\paragraph{Numerical verification.}
Experiment~\texttt{exp\_01\_inertia.py} initialises a Gaussian wavepacket
($\sigma = 3$, $\omega = 0.2$) with momentum $\mathbf{k} = (k,k,k)$,
$k = 0.3/\sqrt{3}$, corresponding to a phase advance
$\theta = \mathbf{k} \cdot \mathbf{V}_1 = 0.3\sqrt{3} \approx 0.520$
rad per hop along the $\mathbf{V}_1 = (1,1,1)$ direction.
Equation~\eqref{eq:v_omega_theta} predicts
\[
    v_x = \tfrac{1}{2}\sin(0.2)\sin(0.3\sqrt{3})
        \approx 0.0493 \text{ nodes/tick}.
\]
The measured x-component fit over 12 ticks gives
$v_x = 0.0464$ nodes/tick — a 6\% discrepancy attributable to
finite packet width and the short run length.  The heavy particle
($\omega = 0.5$, same $\mathbf{k}$) drifts measurably slower,
confirming that $\sin(\omega)$ governs inertial resistance: a larger
phase cost per hop reduces the net kinetic probability imbalance.

% ── Subsection 3 ─────────────────────────────────────────────────────────────
\subsection{Acceleration and Force}
\label{subsec:force}

Acceleration requires a time-varying gradient.  Taking the discrete
time derivative of Eq.~\eqref{eq:v_omega_theta} with $\omega$ held
constant:
\begin{equation}
    a = \frac{\Delta v}{\Delta t}
      = \frac{c}{2}\,\sin(\omega)\,\cos(\theta)\,\frac{\Delta\theta}{\Delta t}.
    \label{eq:acceleration}
\end{equation}
For small gradients ($\cos\theta \approx 1$) this reduces to the
lattice analogue of Newton's second law~\cite{newton1687}:
\begin{equation}
    a = \frac{c}{2m}\,F_\theta,
    \qquad
    F_\theta \equiv \frac{\Delta\theta}{\Delta t}.
    \label{eq:newton}
\end{equation}
The quantity $F_\theta$ — the rate at which the spatial phase gradient
changes per tick — is the lattice definition of force.  A gravitational
or electromagnetic field acts by twisting the phase of the vacuum over
time, unbalancing $P_\mathrm{fwd}$ and $P_\mathrm{bwd}$ and
continuously converting $p_\mathrm{stay}$ into directed kinetic
probability.  The mechanism by which the Coulomb potential produces
$F_\theta$ in the hydrogen orbital is analysed in
Section~\ref{sec:hydrogen}.

% ── Subsection 4 ─────────────────────────────────────────────────────────────
\subsection{The Continuum Limit: Emergence of the Dirac Equation}
\label{subsec:dirac_limit}

The kinematic relations above emerge in the long-wavelength limit $a \to 0$.
Here we show that this limit yields the Dirac equation exactly, with the
gamma matrices identified as geometric encodings of the RGB/CMY sublattice
structure.

\paragraph{The hop operator in Fourier space.}
Working in Fourier space, the hop operator over the RGB sublattice becomes
the structure factor
\begin{equation}
    \tilde{H}_\text{RGB}(\mathbf{k})
        = \frac{1}{3}\sum_{\mathbf{v}\in\text{RGB}} e^{i\mathbf{k}\cdot\mathbf{v}}.
    \label{eq:structure_factor}
\end{equation}
Expanding near $\mathbf{k} = \mathbf{0}$:
\begin{equation}
    \tilde{H}_\text{RGB}(\mathbf{k})
        \approx 1
          + \frac{i}{3}(1,1,-1)\cdot\mathbf{k}
          - \frac{1}{6}\mathbf{k}^T M \mathbf{k}
          + O(k^3),
    \label{eq:hop_expand}
\end{equation}
where $M_{ij} = \sum_{\mathbf{v}\in\text{RGB}} v_i v_j$.
Numerically, $M$ has diagonal entries $3$ and off-diagonal entries $\pm 1$:
\begin{equation}
    M = \sum_{\mathbf{v}\in\text{RGB}} \mathbf{v}\mathbf{v}^T
      = \begin{pmatrix} 3 & -1 & 1 \\ -1 & 3 & 1 \\ 1 & 1 & 3 \end{pmatrix}.
    \label{eq:frame_matrix}
\end{equation}
The off-diagonal terms of $M$ produce cross terms in $\mathbf{k}^T M \mathbf{k}$
that appear to break rotational invariance.
They are cancelled by the CMY sublattice contribution, as shown below.
% NOTE: An earlier draft incorrectly stated M = 3*delta_ij. The correct
% statement is that the full two-sublattice dispersion is isotropic, not
% that M itself is diagonal. See isotropy paragraph below.

\paragraph{Isotropy at the operator level: the lattice is birefringent.}
At the operator level the dispersion is \emph{not} fully isotropic.
The single-tick spinor propagator, derived from the bipartite tick
rule, is
\begin{equation}
    T(\mathbf{k}) = \begin{pmatrix}
        i\sin(\omega/2) & \cos(\omega/2)\,\tilde{H}_\text{RGB}(\mathbf{k}) \\
        \cos(\omega/2)\,\tilde{H}_\text{CMY}(\mathbf{k}) & i\sin(\omega/2)
    \end{pmatrix},
\end{equation}
with $\tilde{H}_\text{CMY}(\mathbf{k}) = \overline{\tilde{H}_\text{RGB}(\mathbf{k})}$
because $\text{CMY} = -\text{RGB}$.  The trace
$\text{tr}\,T = 2i\sin(\omega/2)$ is independent of $\mathbf{k}$, and
the eigenvalues are
$\lambda_\pm(\mathbf{k}) = i\sin(\omega/2) \pm \cos(\omega/2)\,
|\tilde{H}_\text{RGB}(\mathbf{k})|$.
The entire $\mathbf{k}$-dependence enters through
$|\tilde{H}_\text{RGB}|^2$, which to second order is
\begin{equation}
    |\tilde{H}_\text{RGB}(\mathbf{k})|^2
        = 1 - \mathbf{k}^T M_\text{eff}\,\mathbf{k} + O(k^4),
        \quad
    M_\text{eff} = \tfrac{1}{9}
    \begin{pmatrix} 8 & -4 & 4 \\ -4 & 8 & 4 \\ 4 & 4 & 8 \end{pmatrix}
    = -\tfrac{4}{9}\,\mathbb{I} + \tfrac{4}{9}\,M.
    \label{eq:M_eff}
\end{equation}
$M_\text{eff}$ has eigenvalues $\{4/3,\,4/3,\,0\}$, with the zero
eigenvalue along $\mathbf{V}_1+\mathbf{V}_2+\mathbf{V}_3 = (1,1,-1)$.
Direct calculation confirms that for $\mathbf{k}$ along this direction
all three RGB phase advances are equal, so
$|\tilde{H}_\text{RGB}|^2 = 1$ \emph{exactly} at all orders in $\mathbf{k}$.
The lattice therefore exhibits a uniaxial birefringence at the operator
level: along the optical axis $(1,1,-1)$ the dispersion is flat;
in the perpendicular plane the dispersion has coefficient $4/3$.

\paragraph{Recovery of $\mathrm{SO}(3)$ at the observable level.}
The full rotation group $\mathrm{SO}(3)$ is recovered on
$O_h$-averaged observables.  The $\Tdiamond$ lattice has octahedral
point symmetry $O_h$ (order 48), and physical observables must be
$O_h$-invariant.  Averaging the operator-level quadratic form
$\mathbf{k}^T M_\text{eff}\,\mathbf{k}$ over the discrete rotations
in $O_h$ gives
\begin{equation}
    \langle \mathbf{k}^T M_\text{eff}\,\mathbf{k}\rangle_{O_h}
        = \tfrac{\mathrm{tr}(M_\text{eff})}{3}\,|\mathbf{k}|^2
        = \tfrac{8}{9}|\mathbf{k}|^2,
\end{equation}
which is isotropic.  The resulting dispersion relation,
\begin{equation}
    \omega^2(\mathbf{k}) = m^2 + \frac{a^2}{3}|\mathbf{k}|^2 + O(k^4),
    \label{eq:dispersion}
\end{equation}
holds for $O_h$-averaged observables and recovers full $\mathrm{SO}(3)$
in the continuum limit.  The directional gradient
$\tfrac{i}{3}(1,1,-1)\cdot\mathbf{k}$ from the first-order expansion
averages to zero over $O_h$; the residual anisotropy at second order
is the birefringence above.

This is the same mechanism by which staggered fermions in lattice
QCD~\cite{susskind1977,kogutsusskind1975} recover Lorentz invariance
in the continuum limit from a discretisation that breaks it at finite
$a$.  The difference here is that the residual anisotropy is not a
discretisation artifact to be averaged away; it is a closed-form
prediction of the framework, observable as direction-dependent photon
dispersion (Section~\ref{subsec:p7_photon_dispersion}).

\paragraph{The Dirac equation.}
Taking $a \to 0$ with physical mass $m = \omega/a$ held fixed,
Eq.~\eqref{eq:dispersion} recovers the relativistic dispersion relation
$E^2 = m^2 + |\mathbf{p}|^2$ (in units where $c = 1$).
The corresponding equation of motion in position space is
\begin{equation}
    \bigl(i\,\gamma^\mu\,\partial_\mu - m\bigr)\,\Psi = 0,
    \label{eq:dirac}
\end{equation}
where $\Psi = (\psi_R, \psi_L)^T$ is the two-component spinor formed by
the RGB and CMY amplitudes.  The spatial gamma matrices
\begin{equation}
    \gamma^i \sim
    \begin{pmatrix} 0 & \boldsymbol{\sigma}^i \\ \boldsymbol{\sigma}^i & 0 \end{pmatrix}
    \label{eq:gamma_spatial}
\end{equation}
encode the RGB/CMY basis geometry: the Pauli matrices $\boldsymbol{\sigma}^i$
are the projections of the hop directions onto the three spatial axes.
The temporal component
\begin{equation}
    \gamma^0 =
    \begin{pmatrix} I & 0 \\ 0 & -I \end{pmatrix}
    \label{eq:gamma_time}
\end{equation}
reflects the sign flip between even and odd ticks.

\paragraph{Summary of structural consequences.}
The Dirac derivation establishes five results that are geometric
consequences of the bipartite lattice, not additional postulates:
\begin{enumerate}
\item \textbf{Spin-$\tfrac{1}{2}$} is forced by the bipartite structure:
    the spinor has exactly two components because there are exactly two
    sublattices.
\item \textbf{The Clifford algebra} $\{\gamma^\mu,\gamma^\nu\} = 2g^{\mu\nu}$
    holds on $O_h$-averaged observables in the continuum limit.  The
    operator-level frame matrix
    $M = \sum_\text{RGB}\mathbf{v}\mathbf{v}^T$
    (Eq.~\ref{eq:frame_matrix}) has eigenvalues $\{4, 4, 1\}$ rather
    than the diagonal form, so the operator-level anticommutators are
    direction-dependent.  Octahedral averaging gives
    $\langle M\rangle_{O_h} = (\mathrm{tr}\,M/3)\,I = 3I$, recovering
    the standard Clifford anticommutator in the continuum.
\item \textbf{Chirality} ($\psi_R$ vs.\ $\psi_L$) arises from the sign
    flip CMY $= -$RGB: left- and right-handed components are geometrically
    distinct sublattices.
\item \textbf{Rest mass} is the Zitterbewegung oscillation rate:
    $m = \sin(\omega/2)/a$ in the continuum limit.  Mass is not put in
    by hand; it is the frequency at which amplitude oscillates between
    the two sublattices.
\item \textbf{Rotational invariance} holds on $O_h$-averaged
    observables in the continuum limit.  The dispersion relation
    $\omega^2(\mathbf{k}) = m^2 + \tfrac{a^2}{3}|\mathbf{k}|^2$ is the
    $O_h$-averaged form (Eq.~\ref{eq:dispersion}); the residual
    operator-level anisotropy is the lattice birefringence with
    optical axis $\mathbf{V}_1+\mathbf{V}_2+\mathbf{V}_3 = (1,1,-1)$
    (Eq.~\ref{eq:M_eff}, P7 in
    Section~\ref{subsec:p7_photon_dispersion}).
\end{enumerate}
