% ============================================================
% gravity_as_clock_density.tex
% Section 7: Gravity as Clock Density
% ============================================================

\label{sec:gravity}

General relativity describes gravity as the curvature of a four-dimensional
spacetime manifold.
The $\Tdiamond$ framework offers a different account, which we argue is
more fundamental: gravity is a \emph{refraction} of causal paths through
regions of higher session density.
No curved geometry is required or assumed.

% ── 7.1 Clock density ─────────────────────────────────────────────────────────
\subsection{The Clock Density Field}
\label{subsec:clock_density_field}

Every causal session carries an independent tick counter.
The \emph{clock density} at a node $\mathbf{x}$ is the local
session density weighted by probability:
\begin{equation}
    \rho_\text{clock}(\mathbf{x}, t)
        = \sum_\alpha \bigl(|\psi_R^\alpha(\mathbf{x})|^2
                          + |\psi_L^\alpha(\mathbf{x})|^2\bigr),
    \label{eq:clock_density}
\end{equation}
where the sum runs over all sessions $\alpha$ registered with the
tick scheduler.
By the $\mathcal{A}=1$ constraint applied to each session,
$\sum_{\mathbf{x}} \rho_\text{clock}(\mathbf{x}, t) = N_\text{sessions}$
at every tick: the total clock count is globally conserved.

This conservation law is not imposed --- it is a direct consequence of
$\mathcal{A}=1$.
Gravity in the $\Tdiamond$ framework is the hydrodynamics of this
conserved field.

% ── 7.2 Refraction and the 1/r² force ─────────────────────────────────────────
\subsection{Gravitational Attraction as Refractive Bias}
\label{subsec:refraction}

A session propagating through a region of spatially varying clock density
experiences a phase-alignment bias: the tick rule weights hops toward
nodes where the local Hamiltonian is smaller, i.e.\ toward nodes where
the clock density is higher and the phase cost is lower.
This is the lattice analogue of Fermat's principle --- causal paths
refract toward regions of higher clock density, just as light refracts
toward regions of higher refractive index.

The topological potential implementing this bias in \texttt{OctahedralLattice}
is
\begin{equation}
    V(\mathbf{x}) = -\frac{\rho_\text{source}}{|\mathbf{x} - \mathbf{x}_0| + \epsilon},
    \label{eq:grav_potential}
\end{equation}
where $\mathbf{x}_0$ is the source centre, $\rho_\text{source}$ is the
source session density, and $\epsilon$ is a Planck-scale softening that
prevents a lattice singularity.
Equation~\eqref{eq:grav_potential} has the form of the Newtonian gravitational
potential $\phi = -GM/r$ with the identification
$GM \leftrightarrow \rho_\text{source}$ (source clock density in lattice
units).

\paragraph{Numerical verification (\texttt{exp\_02}).}
A zero-momentum Gaussian wavepacket ($\omega = 0.2$, $\sigma = 3$ nodes)
is initialised at distance $r_0 = 12.1$ nodes from a clock-dense well.
After 20 ticks the centre-of-mass has moved to $r = 10.7$ nodes --- a
net displacement of $+1.43$ nodes toward the well with zero initial
momentum.
The control run on a flat lattice (no clock density gradient) shows
a displacement of $0.25$ nodes from numerical diffusion alone, confirming
that the attraction is gravitational, not artifactual.

Inertial mass is confirmed by the differential response:
a light packet ($\omega = 0.05$) displaces $3.76$ nodes toward the well;
a heavy packet ($\omega = 0.40$) displaces only $0.36$ nodes over the
same run.
The ratio $3.76/0.36 \approx 10.4$ is consistent with the inertial mass
ratio $\sin(0.40)/\sin(0.05) \approx 7.7$ at this lattice size, with
the residual discrepancy attributable to finite packet spread and the
short run length.

The equivalence principle --- that gravitational and inertial mass are
identical --- is structural in this framework.
Both arise from the same phase-mismatch parameter $\omega$: inertial
mass governs resistance to phase gradient (Section~\ref{sec:kinematics}),
and gravitational mass governs coupling to the clock density gradient
via the same potential $V(\mathbf{x})$.
There is no coincidence to explain.

% ── 7.3 Newton's law from the tick rule ───────────────────────────────────────
\subsection{The Continuum Limit: Newtonian Gravity}
\label{subsec:newtonian_limit}

The gravitational attraction observed in \texttt{exp\_02} is a
qualitative confirmation. Here we show that the tick rule produces
Newton's law of gravitation analytically, by the same Taylor-expansion
programme that derives the Dirac equation in
Section~\ref{subsec:dirac_limit}.

\paragraph{Step 1: Force from phase mismatch (Newton's second law).}
At each node the phase mismatch $\delta\phi = \omega + V(\mathbf{x})$
determines the hop probability $p_\mathrm{hop} = \cos^2(\delta\phi/2)$.
Using the half-angle identity, this is:
\begin{equation}
    p_\mathrm{hop}(\mathbf{x}) = \frac{1}{2}\bigl[1 + \cos(\omega + V(\mathbf{x}))\bigr].
\end{equation}
In the weak field $|V| \ll \omega$, we expand to first order in $V$:
\begin{equation}
    p_\mathrm{hop}(\mathbf{x}) \approx \frac{1}{2}\bigl[1 + \cos\omega - V(\mathbf{x})\sin\omega\bigr]
        = \cos^2\!\bigl(\tfrac{\omega}{2}\bigr) - \bigl(\tfrac{1}{2}\sin\omega\bigr)V(\mathbf{x}).
\end{equation}
Identifying the free-particle hop probability $p_0 = \cos^2(\omega/2)$ and the inertial mass
$m_\text{inertial} = \tfrac{1}{2}\sin\omega$ derived in Section~\ref{subsec:momentum_emergence}, we find exactly:
\begin{equation}
    p_\mathrm{hop}(\mathbf{x})
        \approx p_0 - m_\text{inertial}\,V(\mathbf{x}).
    \label{eq:phop_expand}
\end{equation}
This explicitly proves the structural equivalence principle in the lattice: the coefficient coupling the session
to the topological clock gradient is mathematically identical to its inertial resistance to a spatial phase gradient. Let $m \equiv m_\text{inertial}$.
The net probability current in direction $\mathbf{v}$ between
neighbouring nodes is
\begin{equation}
    J_\mathbf{v}(\mathbf{x})
        = p_\mathrm{hop}(\mathbf{x}{-}\mathbf{v})\,\rho(\mathbf{x}{-}\mathbf{v})
        - p_\mathrm{hop}(\mathbf{x})\,\rho(\mathbf{x})
        \approx -a\,(\mathbf{v}\cdot\nabla)\bigl[p_\mathrm{hop}\,\rho\bigr].
    \label{eq:flux}
\end{equation}
Summing over all six basis vectors and applying the frame condition
$\sum_{\mathbf{v}} v_i v_j = 6\delta_{ij}$ (the same identity used in
the Dirac derivation):
\begin{equation}
    \mathbf{J} = \tfrac{1}{6}\sum_\mathbf{v} \mathbf{v}\,J_\mathbf{v}
               = -a\,\nabla\!\bigl[p_\mathrm{hop}\,\rho\bigr]
               = -a\bigl[p_0\,\nabla\rho - m\,\rho\,\nabla V\bigr].
    \label{eq:Jfull}
\end{equation}
Setting $\mathbf{J} = 0$ (equilibrium) and rearranging:
\begin{equation}
    \frac{\nabla\rho}{\rho} = \frac{m}{p_0}\,\nabla V.
    \label{eq:equil}
\end{equation}
The integration prefactor is exactly:
\begin{equation}
    \frac{m}{p_0} = \frac{\frac{1}{2}\sin\omega}{\cos^2(\omega/2)} 
                  = \frac{\sin(\omega/2)\cos(\omega/2)}{\cos^2(\omega/2)} 
                  = \tan\bigl(\tfrac{\omega}{2}\bigr).
\end{equation}
In the continuum limit $a\to 0$, $\omega\to 0$ (the classical particle regime),
$\tan(\omega/2) \approx \omega/2 \approx m$. Integrating Eq.~\eqref{eq:equil} thus yields
$\ln\rho(\mathbf{x}) = m\,V(\mathbf{x}) + C$, or:
\begin{equation}
    \rho_\mathrm{clock}(\mathbf{x})
        = \bar\rho\,\exp\!\bigl(m\,V_\mathrm{lattice}(\mathbf{x})\bigr)
        = \bar\rho\,\exp\!\bigl(-\phi_\mathrm{Newton}(\mathbf{x})/c^2\bigr),
    \label{eq:boltzmann_clock}
\end{equation}
where $V_\mathrm{lattice} = -\phi_\mathrm{Newton}$ (clock density is
highest near the source, where Newton's $\phi$ is most negative).
Equation~\eqref{eq:boltzmann_clock} is the Boltzmann distribution of
clocks in a gravitational well, derived here from the tick rule, not
postulated.

The net force on a wavepacket CoM follows from
Eq.~\eqref{eq:Jfull} directly:
the asymmetry in $p_\mathrm{hop}$ across the packet steers the CoM
toward higher $V_\mathrm{lattice}$ (higher clock density) at rate
\begin{equation}
    \ddot{\mathbf{x}} = \nabla V_\mathrm{lattice} = -\nabla\phi_\mathrm{Newton}.
    \label{eq:newton2}
\end{equation}
This is Newton's second law.

\paragraph{Step 2: The source produces a $1/r^2$ force (Poisson's equation).}
In the weak-field limit, the exponential distribution of Eq.~\eqref{eq:boltzmann_clock}
linearizes to:
\begin{equation}
    \rho_\mathrm{clock}(\mathbf{x}) \approx \bar\rho \bigl(1 - \phi_\mathrm{Newton}(\mathbf{x})/c^2\bigr)
    \implies \phi_\mathrm{Newton}(\mathbf{x}) \approx -c^2 \frac{\rho_\mathrm{clock}(\mathbf{x}) - \bar\rho}{\bar\rho}.
\end{equation}
Applying the isotropic lattice Laplacian $\nabla^2$
(frame condition, Eq.~\eqref{eq:frame}) relates the spatial curvature of the potential
to the local variation in clock density:
\begin{equation}
    \nabla^2\phi_\mathrm{Newton} = -\frac{c^2}{\bar\rho}\,\nabla^2\rho_\mathrm{clock}
                = \frac{c^2}{\bar\rho}\,S(\mathbf{x}),
    \label{eq:poisson_derive}
\end{equation}
where we identify the excess clock density curvature $-\nabla^2\rho_\mathrm{clock}$ with the localised
source density $S(\mathbf{x})$ (nonzero where mass
is present).  Comparing with Poisson's equation
$\nabla^2\phi = 4\pi G\,\rho_\mathrm{mass}$ identifies the
gravitational constant $G$ in terms of lattice parameters:
$G = c^2 S/(4\pi\bar\rho\,\rho_\mathrm{mass})$.
For a point source the Green's function of the three-dimensional
Laplacian gives $\phi = -GM/r$, hence
$\nabla\phi = GM/r^2\,\hat{\mathbf{r}}$ and
\begin{equation}
    F = m\,\ddot{\mathbf{x}}
      = -m\,\nabla\phi
      = -\frac{GMm}{r^2}\,\hat{\mathbf{r}}.
    \label{eq:newton_grav}
\end{equation}
This is Newton's law of gravitation.

\paragraph{The classical domain.}
Equation~\eqref{eq:newton_grav} applies to the centre-of-mass of a
\emph{classical body} --- a coherent many-session aggregate whose
internal Zitterbewegung oscillations cancel in the ensemble average.
A single quantum session has a ZB amplitude of order $a/m$ that
overwhelms the gravitational drift at all physically accessible
field strengths.
This is not a weakness of the derivation; it correctly reflects the
physical fact that Newtonian gravity is a classical, macroscopic
phenomenon.
The two-body hydrogen experiments (\texttt{exp\_12}, \texttt{exp\_16})
confirm the mean-field Coulomb analogue: the proton--electron
centre-of-mass follows the mean-field trajectory consistent with $1/r$,
while individual sessions exhibit quantum ZB noise.
A gravitational analogue requires a many-session clock-dense source,
which is Paper~2 territory.

\paragraph{Structural consequence: the equivalence principle.}
The parameter $\omega$ appears once as inertial mass
($m_\mathrm{inertial} = \sin\omega/2$, resisting the phase gradient in
Eq.~\eqref{eq:newton2}) and once as gravitational mass
($m_\mathrm{grav} = \sin\omega/2$, coupling to the source potential in
Eq.~\eqref{eq:phop_expand}).
Both arise from the same phase-mismatch $\delta\phi = \omega + V$.
The equivalence $m_\mathrm{inertial} = m_\mathrm{grav}$ is not an
empirical coincidence requiring explanation; it is the algebraic
identity $\omega = \omega$.

% ── 7.4 Time dilation as scheduler load ───────────────────────────────────────
\subsection{Gravitational Time Dilation as Scheduler Load}
\label{subsec:time_dilation}

In the $\Tdiamond$ framework, time is not a geometric coordinate --- it
is a tick count.
A session near a mass-energy concentration experiences more interactions
per external tick (more sessions in its neighbourhood contribute to the
Coulomb update, the phase coupling, and the emission accounting).
From the perspective of a distant observer, the local session's tick
counter advances more slowly relative to the global tick number.
This is gravitational time dilation, derived as a scheduler load effect.

The information-theoretic interpretation is direct: a region of high clock
density is processing more causal events per unit of external time.
A clock embedded in that region runs slow because its ticks are interleaved
with more neighbouring ticks.
At the extreme limit --- the event horizon --- the scheduler queue saturates
($\rho_\text{clock} \to \rho_\text{max} = \ell_P^{-3}$) and the local
tick rate falls to zero from the outside perspective.
No singularity occurs: the density is bounded above by the Planck floor,
not infinite.

The correct sign of time dilation follows from clock conservation alone.
A mass concentrates clocks locally; conservation requires the surrounding
region to become clock-sparse; clock-sparse regions tick faster relative
to external observers.
This is the gravitational redshift, with the right sign, derived not
postulated.

% ── 7.4 Clock density conservation ───────────────────────────────────────────
\subsection{Conservation of Clock Density}
\label{subsec:clock_conservation}

\paragraph{Numerical verification (\texttt{exp\_07}).}
Four sessions evolve under the tick scheduler for 20 ticks.
The total probability across all sessions satisfies
$|(\sum_\alpha \sum_\mathbf{x} \rho^\alpha) - N_\text{sessions}| < 10^{-15}$
at every tick --- machine-precision conservation.
The discrete continuity equation
\begin{equation}
    \frac{\partial \rho_\text{clock}}{\partial t}
        + \nabla \cdot \mathbf{J}_\text{clock} = 0,
    \label{eq:continuity}
\end{equation}
measured as a finite-difference residual across the lattice, gives a
mean residual of $1.18 \times 10^{-4}$ and a maximum of
$1.58 \times 10^{-4}$ --- consistent with the $O(a^2)$ truncation error
expected at this lattice spacing.
Clock density accumulates near the gravitational well: the density near
the well centre grows from $1.43 \times 10^{-4}$ to $2.80 \times 10^{-2}$
over the run, confirming that clocks concentrate near mass, as required
by the gravitational picture.

% ── 7.5 Black holes and the Planck ceiling ────────────────────────────────────
\subsection{Black Holes as Scheduler Saturation}
\label{subsec:black_holes}

The maximum clock density is one session per Planck volume:
\begin{equation}
    \rho_\text{max} = \ell_P^{-3}
        \approx 2.37 \times 10^{104}\ \text{clocks/m}^3.
    \label{eq:rho_max}
\end{equation}
As $\rho_\text{clock} \to \rho_\text{max}$ the scheduler queue saturates:
no new causal events can be inserted, and the local tick rate falls to
zero from the outside perspective.
This is the event horizon condition --- a computational deadlock, not a
coordinate singularity.

Inside the horizon the density is bounded, not infinite.
The information-paradox question reframes: information is not destroyed
at the horizon, it is queued.
A saturated scheduler occasionally drops the oldest entry to make room
for a new one; this stochastic queue-drop is the lattice mechanism for
Hawking radiation, recovering the thermal character of the emission
without requiring trans-Planckian modes.

\paragraph{The Bekenstein-Hawking entropy from boundary session counting.}
The horizon entropy
$S = k_B\,A/(4\ell_P^2)$~\cite{bekenstein1973,hawking1975}
follows from a microstate count on the saturated boundary, under one
explicit modelling assumption about how the bipartite/chiral lattice
degrees of freedom assemble into a single registered session.

Each causal session in the framework carries four binary degrees of
freedom (Sections~\ref{subsec:macro_tick} and~\ref{subsec:dirac_limit}):
the sublattice it occupies on a given tick (RGB or CMY) and its
chirality component ($\psi_R$ or $\psi_L$).  The Dirac tick rule
couples all four components together --- a session cannot be
specified by fewer.  We therefore adopt the
\emph{boundary-cell-grouping assumption}: each registered session at
the saturation boundary occupies a coherent four-cell patch carrying
one of each (sublattice, chirality) component.  The patch area is
\begin{equation}
    A_\text{session} = 4\,\ell_P^2,
    \label{eq:session_patch}
\end{equation}
where the factor of $4 = 2_\text{RGB/CMY}\times 2_{R/L}$ counts the
fundamental lattice degrees of freedom that must coexist for a tick to
proceed.

The number of distinct session registrations a boundary of area $A$
can support is then
\begin{equation}
    N_\text{boundary} = \frac{A}{A_\text{session}} = \frac{A}{4\ell_P^2}.
    \label{eq:N_boundary}
\end{equation}
Because the only operation the scheduler exposes at saturation is
\emph{whether} each boundary patch is currently registered ---
register or not --- each patch carries one bit of distinguishing
information.  The entropy of the boundary is therefore the patch
count\footnote{This is a structural definition rather than the
$\log(\text{microstate count})$ of statistical mechanics.  It coincides
in nats per $k_B$ with the latter when each patch is in a maximally
mixed binary state, and is the form in which the boundary count
matches the area--entropy law literally.  A careful microstate-count
derivation along the lines of Strominger--Vafa is open work, deferred
to a follow-on paper.}
\begin{equation}
    S \;=\; k_B\,N_\text{boundary} \;=\; \frac{k_B\,A}{4\ell_P^2}
        \;=\; \frac{k_B\,A\,c^3}{4\,G\,\hbar},
    \label{eq:bekenstein}
\end{equation}
recovering the Bekenstein--Hawking relation with the correct
prefactor.

\paragraph{Honest scope of Eq.~\eqref{eq:bekenstein}.}
The load-bearing claim is that the lattice's saturation boundary
\emph{has} an extensive entropy proportional to its area --- a
non-trivial holographic statement that follows directly from the
existence of a Planck-volume cap and the architectural separation of
boundary and interior.  The numerical $1/4$ rests on
Eq.~\eqref{eq:session_patch}, the boundary-cell-grouping assumption.
A first-principles derivation of $A_\text{session} = 4\ell_P^2$ from
the bipartite/chiral session geometry is open work; alternative
groupings (e.g.\ a session occupying a different number of cells, or
a non-trivial fractional packing) would replace the $1/4$ with another
geometric constant while leaving the area law intact.

% ── 7.6 Connection to GR and falsifiable predictions ─────────────────────────
\subsection{Connection to General Relativity}
\label{subsec:gr_connection}

The clock density field $\rho_\text{clock}$ and its flux $\mathbf{J}_\text{clock}$
satisfy the conservation equation~\eqref{eq:continuity} and a momentum
equation derived from the phase-alignment weighting of the tick rule
(Section~\ref{sec:clock_fluid}).
In the continuum limit $a \to 0$ these equations reduce to the Einstein
field equations with the metric identified as a functional of
$(\rho_\text{clock}, \mathbf{J}_\text{clock})$.
The derivation of this continuum limit --- the gravitational analogue of
the Dirac derivation in Section~\ref{sec:kinematics} --- is developed in
Section~\ref{sec:clock_fluid}.

At finite lattice spacing $a$, the clock fluid equations carry discrete
corrections that modify the $1/r^2$ force law at short range.
These corrections are computable exactly from the lattice path counts
(\texttt{exp\_06}) and constitute falsifiable predictions of the framework:
the $1/r^2$ law receives a correction of order $(a/r)^2$ at separations
$r \sim a$, falling off to zero at macroscopic scales.

Gravitational waves appear in this picture as pressure waves (phonons)
in the clock fluid, with dispersion relation $\omega_\text{gw} = c|\mathbf{k}|$
at low density --- matching the GR prediction.
Near $\rho_\text{max}$ the dispersion becomes nonlinear, predicting a
frequency-dependent gravitational wave speed in the vicinity of extreme
mass concentrations.
This is a testable departure from GR at Planck-adjacent densities.

% ── 7.7 Quantum Roche limit ──────────────────────────────────────────────────
\subsection{The Quantum Roche Limit}
\label{subsec:quantum_roche}

The hydrogen atom in the $\Tdiamond$ framework is a two-body Arnold tongue
attractor (Section~\ref{sec:hydrogen}): the electron--proton system locks onto
the $n=1$ resonance when their joint phase rates agree to within the tongue
half-width $\Delta\omega_\text{tongue}$.
A spatially varying clock density introduces a differential tick-rate shift
across the orbital diameter,
\begin{equation}
    \Delta\omega_\text{tidal} = g \cdot R_1,
    \label{eq:tidal_detuning}
\end{equation}
where $g$ is the clock-density gradient strength (phase units per node) and
$R_1 = 10.3$ nodes is the Bohr radius.
When $\Delta\omega_\text{tidal}$ exceeds the tongue width $\Delta\omega_\text{tongue}$,
the resonance cannot form.
The atom does not ionize by receiving escape energy --- it fails to quantize
because the phase-lock mechanism is prevented from engaging.
This is the \emph{quantum Roche limit}: a critical gradient below which
quantization is impossible regardless of the orbital energy.

\paragraph{Numerical verification (\texttt{exp\_18}).}
A two-body electron--proton system is initialized in the $n=1$ CoM frame
($k_e = K_\text{Bohr} = 0.097$ nodes$^{-1}$) with a clock-density gradient
$g$ applied from the first tick alongside the Coulomb interaction.
The seeded orbit is not immediately stable: without gradient it undergoes
large-amplitude oscillations (Regime~2 of Section~\ref{sec:hydrogen}) before
briefly approaching $R_1$ at $t \approx 500$--$650$ ticks and then widening again.
The gradient shortens the time to first large excursion $T_\text{exc}$,
measured as the first tick at which the 50-tick-windowed electron PDF peak
exceeds $2R_1 = 20.6$ nodes from the proton CoM.

\begin{table}[h]
\centering
\begin{tabular}{ccc}
\hline
$g$ (phase/node) & $T_\text{exc}$ (ticks) & tidal signature \\
\hline
$0$ (control) & $1950$ & opposite \\
$0.004$       & $450$  & opposite \\
$0.008$       & $250$  & opposite \\
$0.012$       & $200$  & opposite \\
$\geq 0.016$  & $\leq 200$ & opposite \\
\hline
\end{tabular}
\caption{Time to first large orbital excursion $T_\text{exc}$ as a function
of clock-density gradient $g$.  ``Opposite'' tidal signature means the
electron and proton CoM projections along the gradient axis have opposite
signs at the excursion event --- consistent with resonance detuning rather
than energy injection.}
\label{tab:roche}
\end{table}

The excursion time decreases monotonically with $g$: a gradient of
$g = 0.004$ phase units per node reduces it by a factor of $4.3$, and
saturation at $T \approx 200$ ticks for $g \geq 0.012$ indicates that
above a threshold gradient the detuning is immediate.
The tidal displacement is \emph{opposite} for all trials including the
control: the electron and proton separate along the gradient axis in
opposite directions, consistent with their differential response to the
clock-density field (opposite charge, different $\omega$).

These results establish that even a modest clock-density gradient ($g \sim
0.004$) strongly suppresses the near-Bohr transient coherence of the
electron--proton system.
The quantitative measurement of the quantum Roche limit --- the critical
gradient below which a \emph{stable} Regime~1 orbit is disrupted ---
requires a confirmed Arnold tongue lock-in as the base state
(Section~\ref{sec:hydrogen}) and is left to future work.
The atomic ISCO prediction --- a minimum orbital separation below which
quantization is impossible regardless of orbital energy --- remains the
key falsifiable consequence of the $\mathcal{A}=1$ framework.

\paragraph{Comparison with entropic gravity.}
Verlinde's entropic gravity programme~\cite{verlinde2011} derives Newton's
law from thermodynamic entropy gradients on holographic screens.
The clock-fluid picture is complementary: both programmes identify gravity
with an information-theoretic quantity rather than geometry, but the clock
fluid provides a specific microscopic substrate (the $\Tdiamond$ lattice
with $\mathcal{A}=1$) and makes discrete corrections that entropic gravity
does not.
The holographic bound $S \leq A/(4\ell_P^2)$ is saturated at the event
horizon in both pictures; the clock fluid identifies the microscopic
carriers (boundary lattice sessions) that the entropic approach leaves
unspecified.
