% ============================================================
% introduction.tex
% ============================================================

{\small\textbf{Beta Status (v0.98-RC):} This is a beta release of an ongoing research program. The core framework and the principal derivations are validated; the numerical predictions in Section~\ref{sec:predictions} have been sharpened against current data with explicit modelling assumptions stated where they enter. Two experiments completed in this release (\texttt{exp\_20} and \texttt{exp\_12b}, both \texttt{PART} per Table~\ref{tab:audit}) confirm joint $\mathcal{A}=1$ conservation at machine precision while opening a new question about long-horizon two-body stability that the paper does not yet answer; this and several other follow-on calculations are flagged in the conclusion's \textit{What remains open} subsection. \emph{Status legend used throughout:} \texttt{PASS} = derivation complete or experiment confirms the claim to stated precision; \texttt{PART} = mechanism demonstrated, full quantitative match pending; \texttt{STUB} = analytical result with no numerical confirmation yet (see Table~\ref{tab:audit} for the full status of every claim). Comments and feedback are welcome ahead of the v1.0 release.\\[1em]}

Quantum mechanics and general relativity are the two most precisely
verified theories in the history of science.
Between them they answer, with extraordinary accuracy, every measurable
question about how matter behaves and how spacetime curves: what energies
particles occupy, how they scatter, how they interfere, how massive bodies
attract.
But both leave a different class of questions open ---
why their own foundations hold.

Why are atomic spectra discrete rather than continuous?
The Schr\"odinger equation produces discrete eigenvalues given the right
boundary conditions, but the boundary conditions are loaded into the
formalism before the calculation begins.
Why is the Born rule $|\psi|^2 = \text{probability}$ true?
It is a postulate --- von Neumann's measurement axiom --- not a derivation.
Why do the Dirac gamma matrices have the algebraic structure they do?
Dirac constructed them to satisfy his requirements; they were
not derived from anything geometrically prior.
Why is the gauge group of the Standard Model $\mathrm{SU}(3)\times
\mathrm{SU}(2)\times\mathrm{U}(1)$ and not some other group?
The Standard Model takes its gauge content as given.
And beyond quantum mechanics: why are gravitational and inertial mass
exactly equal?
General relativity declares their equality as the equivalence principle
--- a postulate, not a derivation --- and provides no mechanism for it.

These are not criticisms of quantum mechanics or general relativity.
They are a precise statement of what both theories are silent about,
and that silence has been known since at least
von Neumann~\cite{vonneumann1932} and is uncontroversial.
A framework that answered these questions would not replace quantum
mechanics or general relativity; it would explain them.

\medskip

This paper proposes such a framework.
We show that all five questions have geometric answers, and that all five
answers follow from a single conservation law:

\begin{center}
\textbf{Conservation of probability --- the axiom $\mathcal{A}=1$.}
\end{center}

On a bipartite three-dimensional causal lattice $\Tdiamond$,
each particle is a \emph{causal session}: an independent phase oscillator
carrying amplitude $(\psi_R, \psi_L)$ on the two chirally-opposite
sublattices, subject to $|\psi_R|^2 + |\psi_L|^2 = 1$ at every tick.
No Hamiltonian, no Hilbert space, and no measurement postulate are
assumed.

The framework's dynamics are scale-free: the tick rule and
$\mathcal{A}=1$ constraint operate identically at any lattice spacing
$a$, and all experiments produce dimensionless ratios
(e.g.\ $R_1/a$, $k_\text{Bohr}\cdot a$) that are independent of $a$.
Physical units enter only through a single calibration: identifying
$a = \ell_P$ (one Planck length) and $\tau = t_P$ (one Planck time)
fixes the speed limit $c = a/\tau = \ell_P/t_P$ and converts
dimensionless lattice results into SI predictions.
The numerical experiments in this paper run at arbitrary lattice
spacing; the Planck identification is applied only when connecting
results to physical observables (Section~\ref{sec:substrate}).

The five answers:
\begin{enumerate}
\item \textbf{Quantization is an attractor, not a boundary condition.}
    The discrete mass spectrum arises from Arnold tongue frequency-locking
    between the particle's internal Zitterbewegung frequency and the vacuum
    carrier.
    Any nearby frequency is dynamically pulled into the nearest rational
    resonance basin.
    Discreteness is self-organising, not postulated.

\item \textbf{The Born rule follows from $\mathcal{A}=1$.}
    Probability is amplitude squared because amplitude is the only
    physical quantity on the lattice and $\mathcal{A}=1$ is the
    conservation law that governs it.
    The Born rule is not a measurement axiom; it is a conservation theorem.

\item \textbf{The Dirac gamma matrices encode lattice geometry.}
    The bipartite RGB/CMY sublattice structure \emph{is} the two-component
    Dirac spinor.
    In the continuum limit $a\to 0$, the tick rule converges to the
    Dirac equation $(i\gamma^\mu\partial_\mu - m)\psi = 0$, with the
    gamma matrices identified as the geometric encoding of three pairs
    of antiparallel basis vectors.
    The Clifford algebra $\{\gamma^\mu,\gamma^\nu\} = 2g^{\mu\nu}$ holds
    on $O_h$-averaged observables (the bipartite frame matrix has
    eigenvalues $\{4, 4, 1\}$ and is anisotropic at the operator level;
    its $O_h$-averaged form is diagonal and reproduces the standard
    anticommutator).  The residual operator-level anisotropy along the
    optical axis $(1,1,-1)$ is itself a falsifiable lattice-birefringence
    prediction (P7, P9 in Section~\ref{sec:predictions}).

\item \textbf{Photon emission is a conservation-law necessity.}
    The Coulomb interaction is a probability \emph{source} --- it injects
    amplitude into the electron session from outside.
    $\mathcal{A}=1$ cannot accommodate this within a single session;
    it mandates creation of a new photon session at orbital lock-in.
    Joint momentum conservation across the three-session event,
    \begin{equation*}
      \mathbf{p}_e + \mathbf{p}_p \;=\; \mathbf{p}_e' + \mathbf{p}_p' + \mathbf{p}_\gamma,
    \end{equation*}
    requires the photon's outgoing momentum $\mathbf{p}_\gamma$ to be
    compensated by an opposite shift in the bound-state momentum,
    $\Delta\mathbf{p}_\text{atom} = -\mathbf{p}_\gamma$ --- the photon
    recoil observed routinely in atomic spectroscopy and
    M\"ossbauer-line measurements.
    Emission, recoil, and energy conservation all follow from
    $\mathcal{A}=1$ alone.
    A first numerical implementation (\texttt{exp\_19c}) preserves
    per-session unitarity at all swept emission rates but does not yet
    exhibit settled orbital lock-in; the audit table
    (Table~\ref{tab:audit}) records this row as \texttt{PART}, and a
    controlled three-arm operator comparison (\texttt{exp\_20}) is the
    next implementation step.

\item \textbf{Gravity is refraction, not curvature.}
    Every causal session carries an independent tick counter.
    The local density of these clocks --- the session density weighted by
    probability --- is the gravitational field.
    Causal paths refract toward regions of higher clock density by the
    same mechanism as Fermat's principle in optics: amplitude hops
    preferentially toward nodes where the phase cost is lower.
    No curved geometry is required.
    Gravitational time dilation is scheduler load: a clock near a mass
    processes more causal events per external tick and therefore runs
    slow.
    The equivalence principle is structural: both gravitational and
    inertial mass arise from the same phase-mismatch parameter $\omega$,
    so their equality requires no explanation.
    The event horizon is scheduler saturation --- a computational
    deadlock at the Planck density ceiling $\rho_\text{max} =
    \ell_P^{-3}$ --- with no singularity, and Hawking radiation as
    stochastic queue-drop.
    In the continuum limit the clock fluid equations reduce to the
    Einstein field equations; at finite lattice spacing they carry
    discrete corrections to the $1/r^2$ law that are computable and
    falsifiable.
\end{enumerate}

\medskip

\paragraph{One law at three resolutions.}
The five answers above are not five conservation laws, nor even three
(probability, momentum, energy).  They are three structural resolutions
of a single constraint.  Per session,
$\int |\psi_i|^2\,dx = 1$ is enforced at every tick.  As a joint field
density across all sessions in a bundle,
$\mathcal{A}_\text{joint}(x) = \sum_i |\psi_i(x)|^2$ is the conserved
quantity whose redistribution between sessions implements emission and
annihilation events.  The clock fluid density that sources gravity is
$\mathcal{A}_\text{joint}(x)$ weighted pointwise by per-session
instruction frequency,
\begin{equation*}
  \rho_\phi(x) \;=\; \sum_i \omega_i\,|\psi_i(x)|^2.
\end{equation*}
Local clock-fluid continuity
$\partial_t \rho_\phi + \nabla\!\cdot\!\mathbf{j}_\phi = 0$ is therefore
the same conservation law expressed at field-density resolution with
mass weights, and energy conservation is its $\omega$-weighted spatial
integral.  Probability conservation, mass-energy conservation, and
gravity all share one structural origin.

\medskip

Two further structural results round out the framework.  First, the
electromagnetic gauge action --- which Section~\ref{sec:vacuum_twist}
postulates following the standard lattice-gauge prescription of
Wilson --- can be \emph{induced} from $-\mathrm{Tr}\,\ln D_\text{lat}[U]$ by
integrating out matter (Sakharov / Zeldovich).
Appendix~\ref{sec:induced_gauge_action} carries out the structural
step symbolically: the leading-order term has the Maxwell form on
$O_h$-averaged observables, with a closed-form gauge-sector
birefringence as the operator-level residual.  Only the numerical
$1/g^2$ prefactor remains.  Second, the kinematic, gauge, gravitational,
and (under the recoil-driven Zitterbewegung reading,
Section~\ref{subsec:bipartite_velocity}) inertial-mass anisotropies all
share a single optical axis $(1,1,-1)$.  The predicted concordance
across five independent observation channels is the framework's
strongest empirical claim (P9, Section~\ref{subsec:p9_concordance});
inter-channel disagreement falsifies the inheritance-from-bipartite-geometry
claim, while concordance pins down the cosmological orientation of the
lattice.

\medskip

A further result, which we consider the deepest, concerns the nature of
bound states.
The standard treatment of hydrogen places the proton as a fixed external
potential $V(r) = -e^2/r$ and solves for the electron's eigenstates.
This is not merely an approximation --- it is the wrong physical picture.

A static Coulomb potential is insufficient to produce stable quantised
orbits from the lattice dynamics alone.
The proton must be represented as an active session --- a phase oscillator
participating in the joint $\mathcal{A}=1$ accounting every tick ---
for the electron to explore the well and lock onto resonant orbits
spontaneously.
Hydrogen is not an electron in a field.
It is \emph{two sessions finding a joint Arnold tongue attractor.}

This distinction has directly observable consequences.
In a static well the electron collapses to the potential minimum regardless
of initial momentum; the basin of attraction around the Bohr orbit has
zero width (\texttt{exp\_11} $n=2$: flat landscape across all
$k$-values tested).
With an active proton session, the proton's Zitterbewegung continuously
breaks the spherical symmetry of the potential, providing the stochastic
perturbation that drives the electron into the resonance basin;
the Bohr orbital wavevector is recovered to four significant figures
with no parameter tuning (\texttt{exp\_12}).
The basin of attraction is finite and self-correcting --- the orbit is
stable not because we initialised it exactly, but because the two-session
dynamics continuously restores it.

Photon emission then follows as a three-session event: the joint
proton-electron resonance transitions to a lower Arnold tongue, the
displaced amplitude is absorbed by a new photon session, and the proton
adjusts to the new joint configuration.
The bound-state recoil
$\Delta\mathbf{p}_\text{atom} = -\mathbf{p}_\gamma$ is mandated by
joint momentum conservation, not added as a separate postulate.
A first numerical implementation
(\texttt{exp\_19c}, two-body initialisation with Coulomb-driven
amplitude drain) preserves per-session unitarity at all swept emission
rates but does not exhibit settled orbital lock-in; the audit table
(Table~\ref{tab:audit}) records this row as \texttt{PART}, and a
controlled three-arm operator comparison (\texttt{exp\_20}) is the
next implementation step.

The framework also distinguishes two conditions that standard quantum
mechanics treats as identical.
\emph{Binding} requires only that the total session energy is negative
--- the two sessions remain in each other's neighbourhood.
\emph{Quantization} requires additionally that the joint phase dynamics
lock onto an Arnold tongue attractor at a specific resonant radius.
A proton too light to resist electron recoil can maintain binding ---
the sessions stay together --- but its excessive recoil prevents the
phase dynamics from settling into any resonant configuration.
The system is gravitationally bound but orbitally unquantized.
Standard quantum mechanics has no mechanism for this distinction: a
system is either in an eigenstate or it is not.
The $\Tdiamond$ framework has the mechanism explicitly, and
\texttt{exp\_16} measures it: below a threshold proton mass, the
two-session system occupies a wide, non-resonant orbit rather than
collapsing or quantizing.

\medskip

\paragraph{Relation to prior discrete spacetime programmes.}
The $\Tdiamond$ framework belongs to a tradition of discrete spacetime
proposals that includes causal set theory~\cite{bombelli1987},
'T~Hooft's cellular automaton interpretation~\cite{thooft2016},
Wolfram's computational universe~\cite{wolfram2002}, and Zuse's
Rechnender Raum~\cite{zuse1969}.
It differs from these in the precision of its structural claims and the
number of falsifiable numerical predictions it makes without free
parameters.
The bipartite RGB/CMY frame structure is not a tuning choice; the
frame matrix $M = \sum_\text{RGB}\mathbf{v}_i\mathbf{v}_j^T$ has
eigenvalues $\{4, 4, 1\}$ and a special direction along the optical
axis $\hat{\mathbf{n}}_* = (1, 1, -1)/\sqrt{3}$, and these are
geometric identities of the basis vectors that force both the
Clifford algebra and the dispersion relation $E^2 = m^2 + |\mathbf{p}|^2$
on $O_h$-averaged observables --- with a falsifiable lattice
birefringence along $\hat{\mathbf{n}}_*$ as the residual signature.
The lattice does not approximate known physics --- it derives it from
a single conservation law and a single geometric substrate.

The framework also relates to emergent gravity proposals such as
Verlinde's entropic gravity~\cite{verlinde2011} and Lazarev's
Non-Metric Space Information (NMSI) theory~\cite{lazarev2026},
which also derive gravity from informational structures.
However, these approaches remain phenomenological: they propose
gravity as emergent but do not derive the full Standard Model gauge
structure or provide the detailed lattice dynamics that enable
numerical falsification.
The $\Tdiamond$ framework provides both the derivation and the
computational substrate for testing.

\paragraph{Methodology and the role of $\mathcal{A}=1$.}
The governing philosophy throughout is that $\mathcal{A}=1$ drives every
result --- it is never fitted to a known answer.
The procedure is always the same: write the discrete tick rule, impose
$\mathcal{A}=1$, take the continuum limit, and read off what emerges.
The Dirac equation was not the target of the bipartite tick rule; it is
what fell out when the continuum limit was taken.
The Newtonian $1/r^2$ force law was not inserted into the clock density
field; it is the static limit of the Poisson equation that emerges from
the Taylor expansion of clock density evolution.
The hydrogen Bohr radius was not tuned; the lattice recovered it to four
significant figures with no free parameters (\texttt{exp\_12}).
In every case the test is: run the lattice, do not adjust it to match
the known answer, and report what it produces.
Where the lattice disagrees with known physics at short range --- the
discrete corrections to the $1/r^2$ law, the nonlinear gravitational
wave dispersion near $\rho_\text{max}$, the gradient ionization threshold
--- those disagreements are predictions, not failures.

All numerical experiments are verified against the same criterion:
the result must emerge from genuine lattice dynamics with no post-hoc
parameter adjustment.
Table~\ref{tab:audit} maps every physical phenomenon discussed in the
paper to the lattice mechanism that produces it, the experiment that
confirms it, and its current verification status.

\paragraph{Computational resources.}
All simulations reported here were performed on a single consumer-grade
desktop workstation using standard Python (\texttt{NumPy} array operations)
with no GPU acceleration, no distributed computing, and no supercomputer
allocation.
The largest grids ($65^3 \approx 274\,000$ nodes) complete in seconds to
hours per run.
This is not a limitation of the lattice; it is evidence that the framework
is computationally accessible.
The entire experimental program --- including the two-body hydrogen scan,
the helium ground state, the double-slit interference figure, and the
gravitational tidal ionization probe --- was reproduced without specialised
hardware.
Any result presented here can be independently verified on a modern laptop.

\paragraph{Organisation of the paper.}
Section~\ref{sec:substrate} defines the $\Tdiamond$ lattice, its
bipartite structure, and the calibration of physical units.
Section~\ref{sec:causal_sessions} introduces causal sessions and the
phase oscillator.
Sections~\ref{sec:scheduler}--\ref{sec:kinematics} derive
momentum, force, and the Dirac equation from the tick rule.
Sections~\ref{sec:gravity}--\ref{sec:vacuum_twist} treat gravity
and electromagnetism as distinct phase geometries.
Sections~\ref{sec:interference}--\ref{sec:observer} cover interference,
decoherence, and the observer.
Sections~\ref{sec:predictions}--\ref{sec:hydrogen} present the
falsifiable predictions and the hydrogen results.
Section~\ref{sec:conclusion} states what has been established and
what remains open.

\bigskip
\input{sections/audit_table}
