\section{The Observer as a Clock}
\label{sec:observer}

In standard quantum mechanics the measurement problem is left
philosophically unresolved.  Copenhagen says: before measurement, a
particle is in a superposition; measurement collapses it to a definite
state instantaneously, by an unspecified mechanism, at an unexplained
boundary between quantum and classical~\cite{bohr1928}.  The mechanism
of collapse is not derived from the formalism — it is postulated as an
axiom of interpretation.

The $\unity$ framework replaces this postulate with a precise mechanical
statement: \emph{an observation is the event in which a new clock
session is registered in the \texttt{TickScheduler}}.  There is no
collapse, no special role for consciousness, and no quantum/classical
boundary.  The same tick rule that governs particle propagation governs
the observer.  The asymmetric effect we call ``measurement'' follows
from the asymmetry in coherence length between the two cones involved.

% ── Subsection 1 ──────────────────────────────────────────────────────────
\subsection{Observation as Clock Registration}
\label{subsec:observation}

Every physical system in the framework is a \texttt{CausalSession}: a
normalised spinor $(\psi_R, \psi_L)$ evolving under the Dirac tick rule
on the $\Tdiamond$ lattice, carrying an instruction frequency $\omega$
and a tick counter maintained by the \texttt{TickScheduler}.  When two
sessions are registered in the same scheduler, the scheduler applies
pairwise phase interactions at shared nodes after each tick — the
\emph{clock entanglement} that constitutes observation.

The state space of the scheduler is the set of all orderings of its
registered clocks.  For $n$ clocks this has size $n!$:

\begin{equation}
    |\Omega_n| = n!,
    \label{eq:state_space}
\end{equation}

verified numerically in \texttt{exp\_05}: $1! = 1$, $2! = 2$, $3! = 6$,
\ldots, $6! = 720$, each step monotonically larger than the last.  The
key consequence of Eq.~\eqref{eq:state_space} is that the state space
grows with every new registration and never shrinks: there is no
\texttt{remove\_session} method on \texttt{TickScheduler} by design.
Observation is architecturally irreversible.

The probability constraint $\unity$ is preserved regardless of how the
clocks are ordered.  All three shuffle schemes tested in
\texttt{exp\_05} — sequential, random, and priority — maintain
$|\sum_i \rho_i - n| < 10^{-15}$.  The physics does not depend on
which observer goes first; only the correlations between their futures
depend on the ordering.

% ── Subsection 2 ──────────────────────────────────────────────────────────
\subsection{Asymmetric Phase Exchange and Apparent Collapse}
\label{subsec:apparent_collapse}

When an observer session and a particle session co-exist in the same
scheduler, their cones overlap at shared nodes.  The \texttt{TickScheduler}
applies a phase rotation to both sessions at each overlap, mixing the
phase structures of the two cones.  After the interaction each cone
contains a modified amplitude and phase distribution — not randomly, but
specifically, determined by the phase information that leaked from one
cone into the other.

This is not wavefunction collapse.  Nothing is destroyed.  Each cone
still exists, still satisfies $\unity$, still encodes a full
distribution over possible futures.  What changed is \emph{which futures
are constructively amplified}.

The asymmetry that produces the apparent collapse is a consequence of the
relative coherence lengths of the two cones.  A freshly prepared particle
carries a narrow, precisely structured interference pattern: most of its
probability is concentrated on a small number of paths, and the phase
relationships between those paths are what produces the interference
fringes.  An observer — a macroscopic detector or any sufficiently large
coupled system — carries a broad, incoherent cone: many degrees of
freedom, already deeply mixed, with phases effectively random relative to
each other.

When they overlap, phase information flows in both directions.  But the
effect is asymmetric.  The observer's cone is so large and already so
incoherent that the small phase perturbation from the particle makes
essentially no difference to its future: the observer's possible futures
are barely changed.  The particle's cone is small and precisely coherent.
The phase perturbation from the observer's broad incoherent field is
large relative to the particle's existing phase structure.  The
particle's interference pattern is scrambled significantly — its possible
futures are substantially redistributed.

\texttt{exp\_05} Test~3 demonstrates this directly.  A particle session
($\sigma = 3$, $\omega = 0.2$) propagates freely for 10 ticks and
accumulates a density $\rho_\mathrm{free}$.  The same particle evolving
alongside an observer session ($\sigma = 1.5$, $\omega = 0.5$, broader
frequency, narrower spatial extent) accumulates a density
$\rho_\mathrm{obs}$.  Both satisfy $A = 1$ to machine precision
($|\sum \rho - 1| < 10^{-8}$), but
\[
    \langle |\rho_\mathrm{free} - \rho_\mathrm{obs}| \rangle
    = 1.60 \times 10^{-4},
\]
confirming that the observer altered the particle's future without
violating probability conservation.

\paragraph{The mechanism of apparent collapse.}
Before the observer interaction, the particle's cone encodes a
superposition of futures — say, ``goes left'' and ``goes right'' with
equal amplitude and a specific phase relationship between them.  That
phase relationship is what produces the interference pattern.  After
the observer interaction, the phase relationship between ``goes left''
and ``goes right'' has been scrambled.  The amplitudes are the same,
but the phases are now random relative to each other.  Destructive
interference no longer suppresses either path — the interference
pattern disappears.

From the outside this looks like collapse: the particle ``chose'' left
or right.  What actually happened is that the phase information encoding
the interference between left and right was transferred into the
observer's cone, where it is diluted among the observer's vast number of
degrees of freedom and effectively inaccessible.  The future that was
being suppressed by destructive interference is now equally probable,
not because it was created, but because the information suppressing it
was removed.

The interaction between an observer session and a particle session
modifies the interference structure of the particle's causal cone —
the amplitude distribution encoding the particle's possible futures.
Because the observer's cone carries many more degrees of freedom than
the particle's cone, the phase exchange is asymmetric: the observer's
future is negligibly modified while the particle's future is
substantially restructured.  What appears as wavefunction collapse in
the Copenhagen interpretation is, in the $\unity$ framework, precisely
this asymmetric phase exchange between cones of vastly different
coherence length.  No information is destroyed; it is redistributed
from the particle's coherent cone into the observer's incoherent one,
where it becomes inaccessible rather than absent.

% ── Subsection 3 ──────────────────────────────────────────────────────────
\subsection{Irreversibility and Entropy}
\label{subsec:irreversibility}

The irreversibility of measurement is, in the $\unity$ framework, an
architectural fact rather than a thermodynamic approximation.
\texttt{TickScheduler} has no \texttt{remove\_session} method.  Once a
clock is registered the state space grows from $(n-1)!$ to $n!$, and
this cannot be undone.

This gives a precise lattice definition of entropy increase: the
entropy of the scheduler's state space is $\log |\Omega_n| = \log(n!)$.
Each observation adds one clock and increases the entropy by
$\log n$ — the number of ways the new clock can be interleaved with the
existing $n-1$ clocks.  The second law of thermodynamics, in this
picture, is the statement that observation is irreversible clock
registration: you cannot un-observe a particle.

\paragraph{No special role for consciousness.}
Any physical system that registers as a \texttt{CausalSession} in the
scheduler is an observer.  A photon detector, a dust grain, a passing
cosmic ray — any interaction that entangles two tick counters constitutes
a measurement.  Consciousness is not a prerequisite; it is merely one
instance of a large-cone incoherent system interacting with a
small-cone coherent one.  The difference between a measurement and a
controlled coherent interaction is only one of the structure of the
observer's cone: incoherent cones scramble interference patterns;
carefully prepared coherent cones can impose them.

% ── Subsection 4 ──────────────────────────────────────────────────────────
\subsection{Entanglement as Correlated Cone Structure}
\label{subsec:entanglement}

When two cones overlap and exchange phase in the \texttt{TickScheduler},
they become correlated: the phase structure of each cone now contains
information about the other.  This is entanglement, described exactly
and locally.

Subsequent modification of one cone — by a further observation — propagates
into the correlated phase structure.  The second cone still contains the
phase information from the first exchange, so its future is correlated
with the outcome of the second measurement on the first particle.  This
correlation is not non-local in the sense of faster-than-light signalling.
The correlation was established when the cones overlapped, in a local
interaction bounded by $c$.  The subsequent measurement travels at $c$
within each cone.  No information moves faster than the cone boundary,
but the pre-established correlation means two widely-separated cones can
have correlated futures even without recent interaction — the $\unity$
account of Bell correlations~\cite{bell1964}.
