% ============================================================
% phase_propagation.tex
% Section 5: Phase Propagation and the Unity Constraint
% Corresponds to: src/core/UnityConstraint.py, src/core/CausalSession.py
% Experiments: exp_03_interference.py, exp_05_observer_clock.py
% ============================================================

\label{sec:phase}

Section~\ref{sec:causal_sessions} defined the causal session as a
$U(1)$ oscillator distributed over the bipartite lattice as a
two-component spinor.
Section~\ref{sec:scheduler} placed that session in a multi-session
scheduler.
This section sets out how the distributed oscillator
\emph{propagates}: what the tick operator does to the amplitude at
each macro-tick, how the unity constraint is preserved through
propagation, and how the discrete dynamics reduces to the
Schr\"odinger equation in the long-time limit.

The headline result is that the Schr\"odinger equation is not a
postulate. It is the central-limit-theorem limit of the bipartite
tick rule, derived in two stages: the Dirac equation as the
relativistic continuum limit
(Section~\ref{subsec:dirac_limit}), and the Schr\"odinger equation as
the non-relativistic limit of Dirac. Standard quantum mechanics is a
theorem in the framework, not an axiom.

% ── 5.1 The Tick Operator ────────────────────────────────────────────────────
\subsection{The Tick Operator}
\label{subsec:tick_operator}

One macro-tick on a single session combines three operations: a local
phase advance by $e^{-i\delta\phi(\mathbf{x})}$ at each node, a
directional hop along the basis vectors of the active sublattice, and
a $\unity$-preserving renormalisation. We write the composite
operation as the tick operator $\hat{T}$:
\begin{equation}
    \hat{T}\,\psi(\mathbf{x}) =
        \mathcal{N}\bigl[
            \cos\!\tfrac{\delta\phi}{2}\,\hat{H}\,\psi(\mathbf{x})
          + i\sin\!\tfrac{\delta\phi}{2}\,\psi(\mathbf{x})
        \bigr],
    \label{eq:tick_operator}
\end{equation}
where $\hat{H}$ is the bipartite hop operator (sum over the three
basis vectors of the active sublattice, see
Section~\ref{subsec:macro_tick}), $\delta\phi(\mathbf{x}) = \omega +
V(\mathbf{x})$ is the local phase mismatch, and $\mathcal{N}$ is the
unity-enforcing renormalisation. The $\cos$ and $\sin$ amplitudes
encode the hop/stay split derived in
Section~\ref{subsec:phase_oscillator}: the half-angle structure makes
$\hat{T}$ a $U(1)$ rotation by $\delta\phi/2$ in the plane spanned by
the local node and its neighbours.

$\hat{T}$ is unitary in the continuum limit $a \to 0$. At finite $a$
it has $O(a^2)$ drift away from exact unitarity, and $\mathcal{N}$
corrects the drift. The correction is numerical, not physical: the
discrete operators do not form an exact representation of the $U(1)$
Lie group, but the constraint $\sum_\mathbf{x} \rho(\mathbf{x}) = 1$
\emph{is} the Lie-group requirement that the state remain on $U(1)$,
so $\mathcal{N}$ is the discrete implementation of that constraint
rather than a normalisation hack.

% ── 5.2 Linearity and Superposition ──────────────────────────────────────────
\subsection{Linearity and Superposition}
\label{subsec:linearity}

The tick operator is linear in the spinor amplitude:
\begin{equation}
    \hat{T}\,(\alpha\psi_1 + \beta\psi_2) =
        \alpha\,\hat{T}\psi_1 + \beta\,\hat{T}\psi_2
    \label{eq:linearity}
\end{equation}
for any complex constants $\alpha, \beta$. The renormalisation
$\mathcal{N}$ is non-linear in the global norm but acts equally on
all path contributions; within a tick the propagation is linear, and
the relative phase between paths is preserved exactly.

Linearity is what makes interference possible. When two amplitudes
from different sources arrive at the same node after $N$ ticks they
add as complex numbers (Section~\ref{sec:interference}). The relative
phase determines whether the addition is constructive (bright fringe)
or destructive (dark fringe). A non-linear propagation rule would
mix amplitudes and intensities, and would not exhibit signed
cancellation.

The bipartite tick rule preserves linearity by construction.
Stochastic propagation rules (Markov chains on probability
distributions, for example) do not. The framework is amplitude-based
not probability-based, and the linearity of the tick operator is
what distinguishes it.

% ── 5.3 Phase Routing and the Discrete Hamiltonian ───────────────────────────
\subsection{Phase Routing and the Discrete Hamiltonian}
\label{subsec:phase_routing}

The local Hamiltonian
\begin{equation}
    H(\mathbf{x}) = \omega + V(\mathbf{x})
    \label{eq:local_hamiltonian}
\end{equation}
is a real number at each node --- an element of the $\mathfrak{u}(1)$
Lie algebra (Section~\ref{subsec:phase_oscillator}). The tick
operator advances the local phase by $-iH(\mathbf{x})\,a$ per tick.
In the small-$a$ limit
\begin{equation}
    \hat{T}\,\psi \approx
        \bigl(1 - iH(\mathbf{x})\,a\bigr)\,\psi
        + a\,\hat{D}\,\psi
        + O(a^2),
    \label{eq:tick_smalla}
\end{equation}
where $\hat{D}$ is the discrete spatial derivative implemented by the
hop step. Comparing with the standard time-evolution operator
$1 - iH\,dt + (i/2m)\nabla^2\,dt$ identifies the lattice spacing $a$
with the time step $dt$ and the hop operator with the kinetic term
--- this is the staggered-fermion identification of
Kogut and Susskind~\cite{kogutsusskind1975}.

The discrete Hamiltonian does not act on a vector space of
pre-defined basis states. It is a spatial field of phase rates,
evaluated locally at each node. Different nodes carry different
phase rates, and the tick operator routes amplitude between them
according to $e^{-iH(\mathbf{x})\,a}$. In standard quantum mechanics
this is the Magnus expansion of the time-evolution operator; on the
lattice it is the literal tick rule.

% ── 5.4 The Unity Constraint in Operation ────────────────────────────────────
\subsection{The Unity Constraint in Operation}
\label{subsec:unity_in_operation}

The $\unity$ constraint
\begin{equation}
    \sum_{\mathbf{x}} \bigl(|\psi_R(\mathbf{x})|^2
                             + |\psi_L(\mathbf{x})|^2\bigr) = 1
    \label{eq:unity_propagation}
\end{equation}
is preserved by $\hat{T}$ up to $O(a^2)$. The renormalisation step
$\mathcal{N}$ corrects the drift by dividing both spinor components
by their joint norm. The typical residual at the per-tick level is
$\sim 10^{-15}$ in double precision and scales as $a^2$, so the
correction vanishes in the continuum limit.

The renormalisation has two distinct legitimate uses:

\paragraph{Conservation.}
Per-tick correction of the $O(a^2)$ drift in a single session. This
is the standard usage and is invoked at the end of every macro-tick
in every experiment. The drift is a property of the discrete
operators, not of the physics: the lattice does not allow an exact
representation of the continuous unitary group with finite-step
operators.

\paragraph{Projection.}
After amplitude transfer between sessions (the photon-emission
mechanism of Section~\ref{subsec:emission}), the source session is
projected onto its remaining sub-space and the destination session
receives the transferred amplitude. Each session is renormalised
individually; joint $\unity$ is preserved across the pair. This is
the discrete equivalent of wavefunction collapse onto a sub-space,
implemented as $\unity$ applied per session after the amplitude
budget has changed.

In neither case is $\mathcal{N}$ a free parameter. Both are
applications of the Lie-group constraint
$\sum \rho = 1$ from Section~\ref{sec:causal_sessions}, applied to a
session whose amplitude budget has changed by a defined physical
mechanism (numerical drift, or amplitude transfer to another
session).

% ── 5.5 The Continuum Limit: From Tick Rule to Schrödinger ───────────────────
\subsection{The Continuum Limit: From Tick Rule to Schr\"odinger}
\label{subsec:continuum_to_schrodinger}

Take $a \to 0$ and $\omega \to 0$ with the lattice velocity
$v = a/dt = 1$ and the rest mass $m = \omega/a$ held fixed. The tick
operator becomes
\begin{equation}
    \hat{T} \;\xrightarrow{a \to 0}\;
        e^{-iH\,dt}
        = e^{-i(m + V - \tfrac{1}{2m}\nabla^2 + \cdots)\,dt},
    \label{eq:tick_continuum}
\end{equation}
the standard non-relativistic Schr\"odinger evolution operator. The
$-\tfrac{1}{2m}\nabla^2$ kinetic term is the small-$a$ expansion of
the bipartite hop operator $\hat{D}$, derived in
Section~\ref{subsec:dirac_limit} via the $O_h$-averaged frame
condition $\langle\sum_\mathbf{v} v_i v_j\rangle_{O_h} = 6\delta_{ij}$
(the operator-level form is anisotropic; the operator-level residual
is the lattice birefringence along the optical axis
$\hat{\mathbf{n}}_* = (1,1,-1)/\sqrt{3}$,
Eq.~\ref{eq:optical_axis}).

The full bipartite tick rule reduces to the Dirac equation in the
relativistic limit (Section~\ref{subsec:dirac_limit}); the
non-relativistic limit of the Dirac equation is the Schr\"odinger
equation. So Schr\"odinger emerges as the long-time, large-$N$,
small-$a$ limit of the discrete rule via
\begin{equation*}
    \text{tick rule}
    \;\xrightarrow{a \to 0}\;
    \text{Dirac equation}
    \;\xrightarrow{v \ll c}\;
    \text{Schr\"odinger equation}.
\end{equation*}

This is the right way to think about the framework's relationship to
standard quantum mechanics. The Schr\"odinger equation is not
assumed; it is derived. A measurable departure from
Schr\"odinger predictions in regimes where $N$ is small (high-energy
interferometry, atom interferometers at the Compton wavelength, etc.)
is a falsifiable consequence of the framework
(Section~\ref{sec:predictions}).

% ── 5.6 Boundary Conditions ──────────────────────────────────────────────────
\subsection{Boundary Conditions}
\label{subsec:boundary_conditions}

Lattice experiments require boundary conditions on the spinor field.
The framework supports three kinds, each appropriate to a different
class of experiment.

\paragraph{Periodic.}
The lattice wraps:
$\psi(\mathbf{x} + L\hat{\mu}) = \psi(\mathbf{x})$ for all basis
directions $\hat{\mu}$. This is appropriate for momentum eigenstates
and is used by default in the harmonic experiments
(\texttt{exp\_09}). Periodic boundaries preserve $\unity$ exactly:
the lattice has no edge from which amplitude can escape.

\paragraph{Absorbing.}
A subset of nodes is marked as absorbing: their hop probability is
set to zero, and amplitude arriving at them is removed by the
$\mathcal{N}$ step. This implements barriers (the double-slit
barrier of Section~\ref{subsec:double_slit_setup}) and detector
geometries. Absorbing boundaries do \emph{not} preserve $\unity$
within a single session: per-session probability is reduced as
amplitude enters the absorbed region. Joint $\unity$ across all
sessions is preserved if the absorbed amplitude is transferred to an
external sink session (e.g., a detector).

\paragraph{Reflecting.}
Hops into the boundary node return as hops out. This implements
infinite-potential confinement (a quantum well). $\unity$ is
preserved exactly because no amplitude leaves the session.

\paragraph{Choice of boundary condition.}
The same tick rule operates inside the lattice; the boundary modifies
which hops are allowed at the edge. The choice is part of the
experimental setup, not an additional postulate. Different
experiments use different boundaries: \texttt{exp\_09} (harmonics)
uses periodic; \texttt{exp\_03} (interference) uses absorbing for the
barrier and reflecting for the lattice edge; \texttt{exp\_10}
(hydrogen) uses reflecting for the lattice edge with an interior
Coulomb well. The freedom to choose is the lattice version of the
freedom to choose boundary conditions on a continuum partial
differential equation.

\paragraph{Summary.}
The propagation step is one tick of the bipartite tick rule applied
to the spinor amplitude $(\psi_R, \psi_L)$. It is linear in the
amplitude, preserves $\unity$ up to $O(a^2)$ drift corrected by
renormalisation, and reduces to the Dirac equation in the
relativistic continuum limit and to the Schr\"odinger equation in the
non-relativistic continuum limit. The lattice does not assume
standard quantum mechanics; it derives it.
