% ============================================================
% predictions.tex
% Section 12: Path Counting and Falsifiable Predictions
% Corresponds to: src/experiments/exp_06_path_counting.py
%                 src/utilities/path_counter.py
%                 src/utilities/lattice_calibrator.py
% ============================================================

\label{sec:predictions}

This section is where the paper cashes the falsifiability check it
has been writing throughout. Every prediction here must satisfy four
criteria:

\begin{enumerate}
    \item \textbf{Derived from the lattice rules alone}, with no free
    parameters introduced for the prediction itself.
    \item \textbf{Numerically specific}, giving a number or a
    functional form rather than a qualitative claim.
    \item \textbf{Different from the standard QM/GR prediction} in
    the regime where the lattice substrate is resolved.
    \item \textbf{Measurable in principle} with known or near-future
    technology.
\end{enumerate}

The nine predictions below meet all four criteria. They span
precision atomic spectroscopy, optical-clock comparisons,
high-energy interferometry, anisotropy searches, decay-rate
measurements in gravitational fields, photon-dispersion tests at
short wavelength, and the multi-channel concordance of a single
optical axis (the framework's strongest empirical claim). Each is
connected to a derivation in an earlier section of this paper or
the appendices; this section collects them, gives explicit
numerical scales, and identifies the experimental signatures that
would either support the framework or falsify it.

% ── 12.1 The Calibration Question ────────────────────────────────────────────
\subsection{The Calibration Question}
\label{subsec:calibration_question}

Every numerical prediction in this section depends on a single
parameter: the lattice spacing $a$, the physical length corresponding
to one node-to-node hop. Until $a$ is fixed, the predictions are
functional forms without numerical scales.

Two natural calibrations bracket the plausible range:

\paragraph{Planck calibration ($a \sim \ell_P \approx 1.6 \times 10^{-35}\,\text{m}$).}
The lattice spacing is identified with the Planck length, the scale
at which classical spacetime is expected to break down. Discrete
corrections to standard physics appear only at this scale, well
below current experimental sensitivity for most observables. The
Planck calibration is the conservative choice: it predicts that the
lattice substrate is undetectable by direct experimental probes
short of Planck-scale collisions, and the falsifiable predictions
become observational only in extreme regimes (cosmological
anisotropy, gamma-ray bursts, the vicinity of black holes).

\paragraph{Compton calibration ($a \sim \lambda_C \approx 2.4 \times 10^{-12}\,\text{m}$).}
The lattice spacing is identified with the electron Compton
wavelength. This is the natural scale of the Dirac equation
itself --- the wavelength at which an electron's rest energy
$mc^2$ matches the photon energy $\hbar c/\lambda$. At this
calibration, discrete corrections to standard physics appear at
laboratory scales accessible to precision atomic experiments,
optical-clock comparisons, and high-energy interferometry.

\paragraph{Other calibrations.}
Intermediate scales are possible. The lattice spacing may turn out
to be process-dependent, with effective $a$ varying with the
substrate's local clock density. The bare framework does not fix
$a$; the calibration is what experiment is meant to determine. The
predictions below are stated functionally, with both Planck and
Compton numerical scales given for orientation.

\paragraph{Compton-calibration numbers as upper bounds.}
Prediction P7 below (gamma-ray-burst time-of-flight, see
Section~\ref{subsec:p7_photon_dispersion}) already constrains the
lattice spacing from above: existing observations require
$a \lesssim 10^{-19}\,\text{m}$, which rules out the Compton
calibration ($a \sim 10^{-12}\,\text{m}$) for any prediction whose
signal scales as a positive power of $a$.
The Compton-calibration numbers quoted in the predictions below are
therefore \emph{upper bounds} on the predicted signal strengths ---
the largest values consistent with the framework's matter sector
before P7's constraint is applied.
At the surviving calibration window
$10^{-19}\,\text{m} \gtrsim a \gtrsim \ell_P \approx 10^{-35}\,\text{m}$,
the actual predicted signals are smaller than the Compton-scale
numbers by a factor of order $(a/\lambda_C)^n$, where $n$ is the
power of $a$ in the specific prediction (typically $n = 1$ or $2$).
We retain the Compton numbers throughout this section for two
reasons: they make the functional form concrete, and they identify
the largest plausible signal that any future tightening of $a$ can
yield. The reader should treat them as upper-bound estimates rather
than as the predicted signal strength at the actual lattice
spacing.

% ── 12.2 P1: Discrete Corrections to Interference ────────────────────────────
\subsection{P1: Discrete Corrections to Interference at Small N}
\label{subsec:p1_interference}

\paragraph{Claim.}
The exact lattice path count $P(N, \Delta x, \Delta y, \Delta z)$
deviates from the Gaussian central-limit envelope at small hop
counts $N$. The fringe pattern in the double-slit experiment
(Section~\ref{sec:interference}) acquires discrete corrections that
modify the standard Huygens--Fresnel prediction at high spatial
frequencies and short propagation distances.

\paragraph{Lattice mechanism.}
Section~\ref{subsec:discrete_corrections} derives the form. The
amplitude at a screen node after $N$ ticks is the complex sum over
all lattice paths of length $N$ from source to destination,
weighted by the cumulative tick phase. For $N \gg 1$ the path count
converges to a Gaussian envelope; for $N$ comparable to the lattice
cutoff the discrete combinatorics deviate from the Gaussian by an
amount computable from \texttt{exp\_06}.

\paragraph{Numerical scale (upper bound at Compton calibration).}
At calibration $a$ and propagation distance $L$, the relevant
ratio is $N \sim L/a$:

\begin{itemize}
    \item Planck calibration: $N \sim 10^{34}$ for laboratory $L \sim 1\,\text{m}$;
    discrete corrections far below experimental sensitivity.
    \item Compton calibration (upper bound, see
    Section~\ref{subsec:calibration_question}): $N \sim 10^{12}$ for
    the same $L$; discrete corrections appear at the highest spatial
    frequencies (gamma-ray interferometry, X-ray Bragg edge).
\end{itemize}
At the GRB-constrained $a \lesssim 10^{-19}\,\text{m}$ the predicted
correction is smaller than the Compton-calibration estimate by a
factor of $(a/\lambda_C)^2 \lesssim 10^{-14}$, placing it below the
sensitivity of present interferometry.

\paragraph{Experimental signature.}
The fringe envelope at high spatial frequencies shows a calculable
departure from the Gaussian Fresnel envelope. The departure is
positive at the highest frequencies (the discrete combinatorics
have heavier tails than the Gaussian) and decays as $1/N^2$ at
moderate $N$.

\paragraph{Falsification.}
The functional form is well defined; the testable regime depends on
the calibration. The Compton-scale interferometry test described
above cannot be performed at the surviving calibration window. The
prediction remains observable in principle at the Planck end via
cosmological-baseline gamma-ray interferometry of high-redshift
sources, where $N$ becomes small relative to the propagation
distance.

% ── 12.3 P2: Minimum Time-Dilation Quantum ───────────────────────────────────
\subsection{P2: Minimum Time-Dilation Quantum}
\label{subsec:p2_dilation_quantum}

\paragraph{Claim.}
The minimum increment of gravitational time dilation between two
neighbouring regions is one tick of bookkeeping per registered
clock:
\begin{equation}
    \delta t_\text{min} = t_\text{tick}.
    \label{eq:min_dilation_pred}
\end{equation}
Time dilation is not a continuous function; it is a step function in
units of the tick duration.

\paragraph{Lattice mechanism.}
Section~\ref{subsec:density_dilation} derives this from the
scheduler: each registered session contributes one tick of pairwise
bookkeeping per macro-tick. Between two regions whose clock
densities differ by exactly one session, the local tick rates differ
by exactly one bookkeeping step --- $t_\text{tick}$.

\paragraph{Numerical scale (upper bound at Compton calibration).}

\begin{itemize}
    \item Planck calibration: $t_\text{tick} = t_P \approx 5.4 \times 10^{-44}\,\text{s}$.
    Unmeasurable.
    \item Compton calibration (upper bound, see
    Section~\ref{subsec:calibration_question}): $t_\text{tick} \approx 8 \times 10^{-21}\,\text{s}$.
    Within roughly two orders of magnitude of current optical-clock
    fractional sensitivity ($\sim 10^{-18}$ at millimetre baselines).
\end{itemize}
At the GRB-constrained $a \lesssim 10^{-19}\,\text{m}$, the
corresponding $t_\text{tick}$ is reduced by $a/\lambda_C \lesssim 10^{-7}$,
giving $t_\text{tick} \lesssim 10^{-28}\,\text{s}$ --- far below
current optical-clock sensitivity. The prediction's functional form
remains; its observable scale shifts to whatever near-future
optical-clock technology can resolve.

\paragraph{Experimental signature.}
Optical atomic clocks separated by a calibrated potential gradient
$\Delta\phi$ should resolve a discrete distribution of fractional
offsets when the gradient is sufficiently small. Above the threshold
$\Delta\phi \gtrsim c^2 t_\text{tick}/(t_\text{measurement})$ the
offsets appear continuous; below it the offsets quantise in
multiples of $t_\text{tick}$.

\paragraph{Falsification.}
A continuous distribution of fractional offsets at the smallest
resolvable gradient differences, with no discrete cutoff at any
scale, falsifies the framework's discrete-time substrate at the
corresponding sensitivity. At present sensitivity the Compton
calibration is already inaccessible (P7 rules it out); P2 at the
constrained calibration is not testable with current technology and
becomes a target for future optical-clock improvements.

% ── 12.4 P3: Octahedral Anisotropy ───────────────────────────────────────────
\subsection{P3: Octahedral Anisotropy}
\label{subsec:p3_anisotropy}

\paragraph{Claim.}
The six preferred axes of the bipartite $\Tdiamond$ lattice (the
RGB and CMY basis directions) break perfect rotational isotropy at
sub-lattice scales. Physical observables sensitive to lattice-scale
geometry should show a directional dependence.

\paragraph{Lattice mechanism.}
The frame condition $\sum_\mathbf{v} v_i v_j = 6\delta_{ij}$ used in
the Dirac derivation (Section~\ref{subsec:dirac_limit}) restores
isotropy in the continuum limit. At finite lattice spacing it is
only approximately isotropic: directional terms scale as $a^2$
relative to the leading isotropic term, and survive in observables
that probe scales near $a$.

\paragraph{Numerical scale (upper bound at Compton calibration).}
The leading anisotropy is at order $(a/\lambda)^2$ for any
process at wavelength $\lambda$. At Planck calibration this is
unmeasurable; at Compton calibration it would appear in atomic
spectroscopy at parts-per-trillion sensitivity.
At the GRB-constrained calibration the predicted amplitude is
smaller by $(a/\lambda_C)^2 \lesssim 10^{-14}$ and is therefore
below current sensitivities for atomic measurements.
The CMB anisotropy signature (a multipole pattern with the lattice's
hexapole symmetry) is independent of the lattice spacing in this
sense --- it depends on the bipartite topology of the substrate, not
on $a$ directly --- and remains observable at any calibration if
the framework is correct and the CMB modes are sensitive to the
substrate symmetry.

\paragraph{Experimental signature.}
Two candidate observations:

\begin{itemize}
    \item \textbf{CMB anisotropy patterns.} The six preferred axes
    of the lattice should leave a hexapole signature in the
    cosmic microwave background, distinct from the dipole and
    quadrupole already observed. Existing CMB analyses
    (\cite{planck2020}) are sensitive to multipole structure up to
    high $\ell$; a search for the specific octahedral pattern
    associated with the $\Tdiamond$ basis vectors is a clean test.
    \item \textbf{Atomic-clock comparisons along inequivalent axes.}
    Two clocks at the same height but oriented along an RGB
    direction versus an off-axis direction should disagree by a
    fraction of order $(\lambda_C/L)^2$ for transitions of
    wavelength $\lambda_C$ over baseline $L$.
\end{itemize}

\paragraph{Falsification.}
Demonstrated isotropy of CMB anisotropy beyond the hexapole, or null
results in oriented atomic-clock comparisons at the relevant
sensitivity, falsifies the Compton calibration. P3 is the
speculative member of this section; the lattice mechanism is well
defined but the predicted amplitude is at the edge of what current
technology can resolve.

% ── 12.5 P4: Decay-Rate Clock-Density Correction ─────────────────────────────
\subsection{P4: Decay-Rate Clock-Density Correction}
\label{subsec:p4_decay_correction}

\paragraph{Claim.}
The decay rate of an unstable particle near a mass concentration
acquires a correction beyond the standard special-relativistic time
dilation. The correction is proportional to the local clock-density
gradient and can be of either sign depending on the decay channel.

\paragraph{Lattice mechanism.}
A particle's decay is a tick-counted process: each tick of the
particle's internal clock advances it toward decay with some
probability per tick. In a region of high clock density, the
scheduler has more bookkeeping per macro-tick, and the particle's
clock advances more slowly relative to a distant observer
(Section~\ref{subsec:density_dilation}). The standard special-
relativistic correction $\gamma$ is recovered, but with an
additional shift from the discrete clock-density gradient that does
not appear in continuum SR.

\paragraph{Derivation of the lattice correction.}
The discrete update of the clock-density field, derived from the
all-six-basis-vector Taylor expansion of the hop operator
(Section~\ref{subsec:density_dilation} and the parallel gravity
derivation), reads
\begin{equation}
    \rho_\text{clock}(\mathbf{x},t+1) - \rho_\text{clock}(\mathbf{x},t)
        = \frac{a^2}{6}\,\nabla^2 \rho_\text{clock}(\mathbf{x})
        + S(\mathbf{x}) + O(a^4),
    \label{eq:rho_clock_discrete}
\end{equation}
where the $1/6$ is the frame-condition factor $\sum_\text{ALL} v_i v_j = 6\,\delta_{ij}$
divided over the second-derivative coefficient.  We adopt the
modelling assumption that the muon's per-macro-tick advance rate
inherits this same discrete correction --- i.e.\ that the muon
session's coupling to its local clock-density background is, to
leading order, identical to $\rho_\text{clock}$'s coupling to itself
through Eq.~\eqref{eq:rho_clock_discrete}.  This is a specific
assumption, not a derived result: a full first-principles derivation
would require coupling a single-particle session to the scheduler
load explicitly.

Under this assumption, the fractional shift in the muon's
distant-observer decay rate, beyond the standard SR
$\gamma$ factor, is
\begin{equation}
    \left(\frac{\Delta \tau}{\tau}\right)_\text{lattice}
        = \frac{a^2}{6}\,\frac{\nabla^2 \rho_\text{clock}}{\rho_\text{clock}}
        \approx \frac{a^2}{6}\,\frac{4\pi G\,\rho_\text{mass}}{c^2},
    \label{eq:p4_shift}
\end{equation}
where the second equality uses the Poisson relation $\nabla^2 \phi = 4\pi G\rho_\text{mass}$
and $\rho_\text{clock}/\rho_0 \approx 1 - \phi/c^2$ in the weak-field
limit.

\paragraph{Numerical scale.}
At Earth's surface with mean mass density $\rho_\oplus = 5510\,\text{kg/m}^3$,
the curvature factor evaluates to $4\pi G\rho_\oplus / c^2 \approx 5.1 \times 10^{-23}\,\text{m}^{-2}$.
For a muon ($\tau_\mu \approx 2.2 \times 10^{-6}\,\text{s}$), the
lattice correction beyond the standard gravitational redshift
($\sim 10^{-10}$, well-confirmed by satellite-clock experiments) is
\begin{align}
    \left(\frac{\Delta\tau}{\tau}\right)_\text{lattice}
        &\approx 1.3 \times 10^{-48}
        \quad (\text{Compton calibration},\ a = \lambda_C),
        \nonumber\\
        &\approx 8.6 \times 10^{-62}
        \quad (\text{GRB-bound},\ a \lesssim 10^{-19}\,\text{m}),
        \label{eq:p4_numerical}\\
        &\approx 2.2 \times 10^{-93}
        \quad (\text{Planck},\ a = \ell_P).
        \nonumber
\end{align}
The relevant length scale is the gravitational scale
$L_g \equiv \sqrt{c^2/(4\pi G\rho_\oplus)} \approx 4.4 \times 10^{11}\,\text{m}$,
not Earth's radius: the correction is $(a/L_g)^2 / 6$, and
$L_g \gg r_\oplus$ pushes the magnitude well below any earlier
$(a/r_\oplus)^2$ sketch.

\paragraph{Experimental signature.}
The correction changes sign across mass concentrations and tracks
the local Laplacian of $\rho_\text{clock}$, distinguishing it
in principle from the standard gravitational redshift (which tracks
$\phi$ itself).  The magnitudes in
Eq.~\eqref{eq:p4_numerical}, however, are below any presently
foreseeable muon-lifetime experiment.

\paragraph{Falsification.}
Differential muon-lifetime experiments cannot, at present sensitivities,
falsify the prediction in either direction: the size of the lattice
correction is so far below current precision that null results are
consistent with the prediction.  Falsification would require
either (i) a five-orders-of-magnitude improvement in differential
muon-lifetime sensitivity beyond the foreseeable, or (ii) the
detection of a clock-density-gradient residual much larger than
Eq.~\eqref{eq:p4_numerical} predicts --- which would falsify the
modelling assumption (single-session inheritance of the
$\frac{a^2}{6}\nabla^2$ correction) rather than the framework's
qualitative claim of a clock-density correction.

% ── 12.6 P5: Arnold Tongue Width — Stark/Lamb Discontinuity ──────────────────
\subsection{P5: Arnold Tongue Width and the Stark/Lamb Discontinuity}
\label{subsec:p5_tongue_width}

\paragraph{Claim.}
Atomic energy levels in this framework correspond to Arnold tongues
of finite width in $(\omega, V)$ parameter space
(Section~\ref{subsec:tongue_width_predictions}). An external
electric field of magnitude proportional to the tongue half-width
should produce a \emph{discontinuous} loss of quantisation, in
contrast to QED's smooth perturbative Stark broadening.

\paragraph{Lattice mechanism.}
Section~\ref{subsec:arnold_tongues} establishes that quantised
states are 2D regions of parameter space, not 1D lines. Inside the
tongue the resonance is robust; at the boundary it breaks; beyond
it the system enters Regime~2 of Section~\ref{subsec:exp16} (bound
but unquantised). An external perturbation displaces the operating
point in $(\omega, V)$; once it crosses the tongue boundary, the
spectroscopic line vanishes.

\paragraph{Numerical scale.}
We extract the tongue half-width from the high-resolution harmonic
scan of \texttt{exp\_09c} (data file
\texttt{exp\_harmonic\_hires\_powermap.npy}) by tracking the fraction
of spectral power that concentrates at the locked frequency
$f = 1/3$ as the instruction frequency $\omega$ is swept across
$\omega_e = \pi/3$.  The full-width-at-half-maximum of the resulting
$P(\omega)$ profile, with linear interpolation between adjacent
$\omega$ rows, gives
\begin{equation}
    \frac{\Delta\omega_\text{tongue}}{\omega_e}
        = 0.033 \pm 0.020
    \quad (3{:}1\ \text{resonance}),
    \label{eq:tongue_width_measured}
\end{equation}
where the uncertainty combines the row-to-row $\omega$ quantisation
($\pi/302$) with the sensitivity to the analysis bandwidth and
smoothing window in quadrature.  The extraction script is
\texttt{src/utilities/tongue\_width\_3to1.py}; the central estimate
agrees with the visible width of the 3:1 tongue in
Figure~\ref{fig:harmonic_pair} and supersedes the earlier eyeball
estimate of $\sim 0.05$.  A dedicated longer-baseline scan would
reduce the analysis-systematic component of the uncertainty.

\paragraph{Critical field from the Peierls substitution.}
A static external field $\mathbf{E} = E\,\hat{\mathbf{x}}$ enters the
tick rule through the scalar-potential gauge $\varphi(x) = -E x$ of
the Peierls coupling derived in
Section~\ref{subsec:em_matter_side} (the gauge-equivalent vector
form $A_x(t) = -E\,t$ gives the same physics).  In this gauge the
local phase advancement at site $x$ becomes
\begin{equation}
    \delta\phi(x) = \omega_e + V_\text{Coulomb}(x) + \frac{e E}{\hbar}\,x,
    \label{eq:stark_detune}
\end{equation}
i.e.\ the field detunes the tick rate by $eE\,x/\hbar$, linear in
position.  Across an orbit of radius $R_1$ the spread of detunings is
$\bigl(eE/\hbar\bigr)\!\cdot\!2 R_1$.  The 3:1 joint resonance breaks
when this spread exceeds the tongue full-width
$2\,\Delta\omega_\text{tongue}$, giving
\begin{equation}
    E_\text{crit}^{(n=1)}
        = \frac{\hbar\,\Delta\omega_\text{tongue}}{e\,R_1}.
    \label{eq:E_crit}
\end{equation}
(Numerical factors of order unity from orbital averaging are absorbed
in the order-of-magnitude estimate below.)  Substituting the
measurement of Eq.~\eqref{eq:tongue_width_measured} with the physical
hydrogen ground-state parameters $\hbar\omega_e = E_{n=1} = 13.6$\,eV
and $R_1 = a_0 = 0.529$\,\AA{} gives
\begin{equation}
    E_\text{crit}^{(n=1)}
        = \frac{0.033\,E_{n=1}}{e\,a_0}
        \approx 8.5 \times 10^{9}\,\text{V/m},
    \label{eq:E_crit_value}
\end{equation}
with the propagated $\pm 0.020$ uncertainty in
$\Delta\omega_\text{tongue}/\omega_e$ giving a range $3 \times 10^9$
to $1.4 \times 10^{10}$\,V/m.  The result is calibration-independent
in the continuum limit: the dimensionless ratio
$\Delta\omega_\text{tongue}/\omega_e$ is set by the resonance
structure of the lattice tick rule and survives the limit
$a/a_0 \to 0$, while $E_{n=1}$ and $a_0$ are the physical
ground-state observables of hydrogen.

A field of $\sim 10^{10}$\,V/m is in principle within reach of
laser-induced static fields in trapped-ion spectroscopy.  The
qualitative signature --- a discontinuous threshold rather than
smooth Stark broadening --- is the load-bearing claim, but
Eq.~\eqref{eq:E_crit_value} now sets a specific numerical target.

\paragraph{Experimental signature.}
Trap a single hydrogen atom in a calibrated DC electric field of
adjustable magnitude. Sweep the field through $\mathbf{E}_\text{crit}$.
QED predicts smooth quadratic Stark broadening; the lattice
predicts a sharp threshold at which the line abruptly broadens or
disappears entirely. The transition should be discontinuous
(within experimental resolution), not smooth.

\paragraph{Falsification.}
The qualitative signature --- a discontinuous threshold at any
finite field magnitude rather than smooth quadratic Stark broadening
to arbitrarily high precision --- falsifies the framework's
tongue-width picture if absent. The numerical threshold depends on
the calibration; the existence of \emph{any} threshold is the
load-bearing claim. This is the most testable near-term prediction
in the paper: the experimental geometry is standard (DC fields on
trapped hydrogen) and the qualitative signature (smooth vs.\
discontinuous broadening) is unambiguous regardless of where the
threshold sits.

% ── 12.7 P6: Emission Rate Corrections Near Mass ─────────────────────────────
\subsection{P6: Emission-Rate Corrections Near Mass Concentrations}
\label{subsec:p6_emission_correction}

\paragraph{Claim.}
Spontaneous emission rates near massive bodies acquire two
corrections beyond the standard gravitational redshift of the
emitted photon's frequency: a clock-density correction to the
electron--proton joint resonance, and a finite-lattice correction at
high frequency.

\paragraph{Lattice mechanism.}
Section~\ref{subsec:emission} derives spontaneous emission as a
forced three-session event when the joint Arnold tongue resonance
jumps. The rate at which tongue-jumps occur depends on the local
clock density (which detunes the joint resonance,
Section~\ref{subsec:density_dilation}) and on the lattice spacing
relative to the photon wavelength (which modifies the available
phase space for the new photon session).

\paragraph{Numerical scale (order-of-magnitude estimates).}
For a hydrogen atom at distance $r$ from a body of mass $M$:

\begin{itemize}
    \item \textbf{Clock-density correction:}
    $\Delta\Gamma/\Gamma \sim (GM/c^2 r) \cdot (\Delta\omega_\text{tongue}/\omega_e)$.
    With the measurement
    $\Delta\omega_\text{tongue}/\omega_e = 0.033 \pm 0.020$ from P5
    (Eq.~\ref{eq:tongue_width_measured}) and the gravitational
    potential of Earth's surface giving $GM/c^2 r \sim 10^{-9}$, the
    correction is of order $3 \times 10^{-11}$ for Lyman-$\alpha$ ---
    within an order of magnitude of the gravitational redshift but
    with a different $r$-dependence.
    \item \textbf{Finite-lattice correction:}
    $\Delta\Gamma/\Gamma \sim (a/\lambda_\gamma)^2$, vanishing at
    Planck calibration, of order $10^{-12}$ at Compton calibration
    (upper bound) for visible-light transitions. At the
    GRB-constrained calibration this is $\lesssim 10^{-26}$ for
    visible light, potentially observable at the highest emission
    frequencies (X-ray and gamma-ray) where the ratio is closer to
    unity.
\end{itemize}

\paragraph{Experimental signature.}
Compare emission rates of identical atoms in spacecraft at varying
gravitational potentials, and across emission-line frequencies
spanning radio to UV. The clock-density correction depends on
$r^{-1}$; the finite-lattice correction depends on
$\lambda_\gamma^{-2}$. The two are independently extractable from a
2D scan over $(r, \lambda_\gamma)$.

\paragraph{Falsification.}
Confirmation of the standard QED emission rate to fractional
precision $10^{-12}$ across the entire $(r, \lambda_\gamma)$ scan,
with no clock-density or wavelength dependence beyond the redshift,
falsifies the Compton calibration.

% ── 12.8 P7: Photon Dispersion at the Lattice Cutoff ─────────────────────────
\subsection{P7: Photon Dispersion at the Lattice Cutoff}
\label{subsec:p7_photon_dispersion}

\paragraph{Claim.}
Photon group velocity decreases at lattice-scale wavelengths,
following the dispersion relation
\begin{equation}
    \omega_\gamma(\mathbf{k})
        = c\,\frac{\sin(|\mathbf{k}|\,a/2)}{a/2},
    \label{eq:photon_disp_pred}
\end{equation}
which goes to zero at the Brillouin zone boundary
$|\mathbf{k}| = \pi/a$.

\paragraph{Lattice mechanism.}
Section~\ref{subsec:brillouin_zone} derives Eq.~\eqref{eq:photon_disp_pred}
directly from the bipartite tick rule for $\omega = 0$ massless
sessions. The dispersion is exact on the lattice, not a perturbative
approximation. \texttt{exp\_09} Part~C measures the linear regime
and confirms it.

\paragraph{Numerical scale.}

\begin{itemize}
    \item Planck calibration: deviation from $\omega = ck$ becomes
    1\% at $\omega \approx 10^{42}\,\text{Hz}$ (unmeasurable directly,
    accessible only through cosmological gamma-ray observations of
    high-redshift sources).
    \item Compton calibration: 1\% deviation at $\omega \approx 10^{18}\,\text{Hz}$
    (gamma-ray interferometry).
\end{itemize}

\paragraph{Experimental signature.}
Time-of-flight measurements of cosmological gamma-ray bursts: high-
energy photons should arrive systematically later than low-energy
photons from the same burst, by an amount proportional to the source
distance and to the squared photon frequency. Existing limits from
GRB observations (\cite{vasileiou2013}) are
$|a| \lesssim 7.6 \times 10^{-19}\,\text{m}$, ruling out the
Compton calibration at this prediction and constraining the lattice
spacing to be substantially below $\lambda_C$.

\paragraph{Falsification.}
The Compton calibration is already falsified for P7 by the GRB
constraints above. The framework survives only at calibrations with
$a \lesssim 10^{-19}\,\text{m}$ for this specific prediction; the
Planck calibration ($a \sim 10^{-35}\,\text{m}$) is consistent with
all current GRB observations. P7 therefore narrows the calibration
range from above: any working calibration must satisfy
$\ell_P \le a \lesssim 10^{-19}\,\text{m}$.

\paragraph{Direction-resolved refinement (lattice birefringence).}
The bipartite RGB/CMY structure produces an anisotropy in the operator-
level dispersion (Section~\ref{sec:kinematics}, Eq.~\eqref{eq:M_eff}):
the spinor propagator's eigenvalue magnitude depends on $\mathbf{k}^T
M_\text{eff}\,\mathbf{k}$, with $M_\text{eff}$ having eigenvalues
$\{4/3,\,4/3,\,0\}$ and the zero eigenvalue along
$\mathbf{V}_1+\mathbf{V}_2+\mathbf{V}_3 = (1,1,-1)$.  Photons
propagating along this optical axis experience the operator-level
flat-dispersion regime; photons propagating in the perpendicular
plane experience the standard quadratic suppression with a
direction-dependent coefficient.  In standard birefringent-crystal
language, the lattice is uniaxial.

The refinement is observable as a direction-resolved bound on the
GRB time-of-flight constraint above.  Existing analyses
(\cite{vasileiou2013}) integrate over sky direction and yield the
isotropic bound $|a| \lesssim 7.6\times 10^{-19}\,\mathrm{m}$.  A
direction-resolved binning of the same data — separating GRBs by
angle relative to a candidate cosmic optical axis — should reveal
that bursts close to $(1,1,-1)$ exhibit \emph{less} dispersion than
perpendicular bursts.  The detection signature is a sharp axial
contrast: the same source, viewed along different sky directions,
yields different dispersion residuals.  The non-detection signature
is a tighter direction-resolved bound on $a$ in every bin.

This is the canonical falsifying signature of lattice birefringence:
**a single light pulse propagating oblique to the optical axis
separates into two arrival peaks** at large baselines, because the
two propagator eigenmodes $\lambda_\pm$ have equal magnitude but
different phases (hence different group velocities).  At
cosmological distances and Compton-bound lattice spacings the
splitting magnitude is order $(a/\lambda_\gamma)^2$ times the
total time-of-flight, putting it at the edge of current
GRB and CMB polarisation precision.

% ── 12.9 The Quantum Roche Limit ─────────────────────────────────────────────
\subsection{P8: The Quantum Roche Limit (Atomic ISCO)}
\label{subsec:p8_quantum_roche}

The eighth prediction of the framework is the quantum Roche limit
derived in Section~\ref{subsec:quantum_roche}: a critical
gravitational gradient below which atomic quantisation is impossible
regardless of orbital energy. We do not repeat the derivation here
but flag it as the eighth member of the predictions set.

\paragraph{Claim, brief.}
A gravitational tidal gradient $g$ across an atom of orbital radius
$R_n$ produces a differential clock-rate shift
$\Delta\omega_\text{tidal} = g R_n$. When this exceeds the tongue
width $\Delta\omega_\text{tongue}$, the joint resonance cannot
form. Below the corresponding critical gradient atomic spectra are
suppressed entirely, not merely shifted.

\paragraph{Numerical scale and signature.}
Critical gradient: $g_\text{crit} \sim \Delta\omega_\text{tongue}/R_n$.
Observable in the vicinity of neutron stars and black holes: the
hydrogen Lyman-$\alpha$ line should be absent in the inner
photosphere of a high-mass neutron star, with an outer cutoff at
the radius where $g = g_\text{crit}$. The signature is a
\emph{gap} in the atomic spectrum across the Roche radius, not a
broadening.

% ── 12.10 P9: Multi-channel concordance of the optical axis ─────────────────
\subsection{P9: Multi-Channel Concordance of the Optical Axis}
\label{subsec:p9_concordance}

\paragraph{Claim.}
The kinematic, gauge, gravitational, and (under the recoil-driven
Zitterbewegung hypothesis of Section~\ref{subsec:bipartite_velocity})
inertial-mass anisotropies all share a single optical axis
$\mathbf{V}_1 + \mathbf{V}_2 + \mathbf{V}_3 = (1, 1, -1)$ in lattice
coordinates.  Independent observational channels probing different
physical phenomena should yield concordant directions when projected
onto sky coordinates.  This is the framework's strongest empirical
claim: one geometric axis, multiple independent witnesses.

\paragraph{Lattice mechanism.}
Each anisotropy inherits its eigenstructure from the bipartite RGB/CMY
frame matrix $M$ (Eq.~\ref{eq:frame_matrix}).  The kinematic-sector
birefringence is governed by $M_\text{eff}$ (Eq.~\ref{eq:M_eff},
eigenvalues $\{4/3, 4/3, 0\}$, optical axis along $(1,1,-1)$); the
gauge-sector birefringence by the bipartite plaquette tensor
$\mathbf{Q}$ (Appendix~\ref{sec:induced_gauge_action}, eigenvalues
$\{4, 4, 16\}$, same axis after Hodge dualising
F-space to B-space, Eq.~\ref{eq:optical_axis_recovered});
gravitational birefringence inherits the same axis through the tilted
Laplacian of Section~\ref{sec:gravity}; and direction-dependent rest
mass under recoil-driven Zitterbewegung inherits the axis through the
same channel structure.  $\mathrm{SO}(3)$ is recovered only on
$O_h$-averaged observables in all four sectors; the residual
operator-level anisotropy is the optical axis.

\paragraph{The five observational channels.}
\begin{enumerate}
    \item \textbf{Kinematic photon dispersion} (P7,
        Section~\ref{subsec:p7_photon_dispersion}).  GRB time-of-flight
        residuals binned by sky direction relative to a candidate
        cosmic optical axis.  Bursts close to the axis show
        \emph{less} dispersion than perpendicular bursts.
    \item \textbf{Vacuum (gauge-sector) birefringence}
        (Appendix~\ref{sec:induced_gauge_action}).
        Polarisation rotation of GRB / CMB photons aligned with the
        same axis.  Constrained at the few-degrees level over Hubble
        baselines by Planck~\cite{planck2020}; direction-resolved
        re-binning is the test.
    \item \textbf{Gravitational birefringence.}  Gravity inherits the
        same eigenstructure (Section~\ref{sec:gravity}); LIGO/Virgo
        polarisation-mode timing should split direction-dependently
        along the same axis.  Existing public data covers it.
    \item \textbf{Direction-dependent rest mass}
        (Section~\ref{subsec:bipartite_velocity}, recoil hypothesis).
        Atomic-clock comparisons in different lab orientations should
        show systematic offsets correlated with the cosmic projection
        of $(1,1,-1)$.  At sub-leading scale
        $\sim (a/\lambda_C)^2 \lesssim 10^{-14}$ at the GRB-bound
        calibration; at the edge of current optical-clock precision
        (parts in $10^{17}$).
    \item \textbf{Anomalous-dispersion anisotropy near atomic
        resonances} (P3 + Lorentz-oscillator interpretation).  Same
        atomic-clock geometry, applied at frequencies near
        spectroscopic resonance lines.
\end{enumerate}

The concordance prediction is sharp: any single channel detecting a
preferred direction constrains the others.  All five must agree on
which way the axis points.

\paragraph{Time-dependence: three scenarios.}
The framework predicts the axis exists in lattice coordinates but does
not fix the embedding into observable sky coordinates --- that is set
by cosmological initial conditions.  Three viable scenarios:
\begin{itemize}
    \item \textbf{Uniform embedding.}  Single global axis, fixed at
        the Big Bang and constant everywhere and forever.  All
        channels at all epochs agree.
    \item \textbf{Domain structure.}  The lattice has ``grains'' with
        different orientations; light passing through different
        grains sees different effective axes.  Spatially varying but
        locally time-stationary.
    \item \textbf{Dynamical embedding.}  The axis at CMB emission
        ($z \sim 1100$) differs from today.  Observations of different
        epochs would show the axis at different phases of evolution.
\end{itemize}
Two empirical handles distinguish them.  The \emph{fast} handle is
sidereal modulation in atomic-clock data: as Earth rotates and orbits
through the cosmic lattice frame, atomic-clock anisotropy should
modulate at sidereal and annual frequencies if the axis is locally
stationary.  The \emph{slow} handle is redshift-resolved binning: GRB
dispersion residuals binned by source redshift would reveal whether
the axis at $z \sim 2$ matches the axis at $z \sim 0.1$.  A persistent
offset is the signature of dynamical embedding.

Under the recoil-driven Zitterbewegung reading
(Section~\ref{subsec:bipartite_velocity}), dynamical embedding is also
the empirical handle on the question of whether the universe is
self-contained or driven from outside the lattice: an externally-driven
system has reason for its embedding to drift, while a self-projecting
one does not.

\paragraph{Numerical scale.}
Channel-dependent.  At the GRB-bound calibration
$a \lesssim 10^{-19}\,\text{m}$:
\begin{itemize}
    \item Channels (1)--(2) (kinematic / gauge): photon dispersion
        anisotropy $\sim (a/\lambda_\gamma)^2$, putting the
        direction-resolved bound at the edge of current GRB / CMB
        polarisation precision.
    \item Channel (3) (gravitational): polarisation-mode timing
        anisotropy $\sim (a/\lambda_\text{GW})^2$; well below current
        LIGO/Virgo sensitivity but tightens under future detector
        improvements.
    \item Channels (4)--(5) (atomic-clock / Zitterbewegung-recoil):
        rest-mass anisotropy $\sim (a/\lambda_C)^2 \lesssim 10^{-14}$
        at the second-order scale; at the edge of current
        atomic-clock precision (parts in $10^{17}$).
\end{itemize}
A signal at the leading $a/\lambda$ rate ($\sim 10^{-7}$) in any
channel is already excluded by existing isotropic bounds; the
prediction is at the second-order direction-resolved scale.

\paragraph{Experimental signature.}
\textbf{Detection:} any two of the five channels finding a preferred
direction that agree (within sky-localisation precision).  The
signature is concordance: independent physical phenomena pointing to
the same axis.  The sharpest version is GRB time-of-flight (channel
1) plus CMB polarisation rotation (channel 2) plus LIGO/Virgo timing
splits (channel 3) all converging on a single sky direction.

\textbf{Within-channel structure:}  the kinematic and gauge sectors
predict \emph{dual} ellipsoid shapes (kinematic prolate, gauge
oblate; Figure~\ref{fig:induced_action_ellipsoid}) on the same axis,
so within each channel the sign of the perpendicular-vs-axial contrast
is itself a sub-prediction.  Mixing up which direction shows
\emph{more} dispersion versus \emph{more} polarisation rotation is a
sub-falsification.

\paragraph{Falsification.}
Three modes, each sharp:
\begin{itemize}
    \item \textbf{Inter-channel disagreement.} Two channels detecting
        a preferred direction that point to different sky locations
        falsifies the inheritance-from-bipartite-geometry claim.
        Channels (1)--(5) are predicted to align; if they do not,
        the framework's unified-anisotropy structure is wrong.
    \item \textbf{Wrong dual structure.} A channel detecting a
        direction with the perpendicular-vs-axial sign opposite to
        what the dual-ellipsoid analysis predicts (kinematic flat
        along axis vs.\ gauge steep along axis) falsifies the
        appendix's induced-action structural derivation.
    \item \textbf{Comprehensive null at predicted sensitivity.} No
        preferred direction in any of the five channels at the
        $(a/\lambda)^2$-level sensitivity does not formally falsify
        the framework, but it tightens $a$ further beyond the current
        $10^{-19}\,\text{m}$ bound and erodes the framework's
        empirical falsifiability.  At sufficient null sensitivity the
        framework becomes consistent with all data but unobservably so
        --- the same status as a Planck-calibrated theory.
\end{itemize}

P9 is therefore best read as the meta-prediction that ties the
others.  P3, P7, the gauge-sector birefringence in
Appendix~\ref{sec:induced_gauge_action}, and the recoil-driven
Zitterbewegung anisotropy of Section~\ref{subsec:bipartite_velocity}
are not independent claims; they are five views of the same single
geometric axis.

% ── 12.11 Summary Table ──────────────────────────────────────────────────────
\subsection{Summary Table of Predictions}
\label{subsec:predictions_summary}

\begin{center}
\renewcommand{\arraystretch}{1.20}
\begin{tabular}{p{0.55in}p{1.6in}p{1.6in}p{1.4in}}
\hline
\textbf{Prediction} & \textbf{Signature} & \textbf{Compton-cal. upper bound} & \textbf{Test} \\
\hline
P1 & Discrete fringe correction at small $N$ & $\sim 10^{-12}$ at gamma-ray $L \sim 1\,\text{m}$ & X-ray Bragg / GRB interferometry \\
P2 & Quantised time dilation $\delta t = t_\text{tick}$ & $t_\text{tick} \sim 8 \times 10^{-21}\,\text{s}$ & Optical-clock comparison at mm baseline \\
P3 & Octahedral CMB hexapole + atomic anisotropy & $(\lambda_C/L)^2 \sim 10^{-12}$ atoms; CMB visible & Planck multipole search; oriented clocks \\
P4 & Decay-rate residual beyond SR $\gamma$ & $\sim 10^{-48}$ for muons at Earth's surface & Differential muon lifetime (currently unfalsifiable) \\
P5 & Discontinuous Stark threshold & $\mathbf{E}_\text{crit} \sim 10^9\,\text{V/m}$ for hydrogen $n=1$ & Trapped-ion DC-field spectroscopy \\
P6 & Emission-rate $r,\lambda$ dependence & $10^{-10}$ in $r$, $10^{-12}$ in $\lambda$ & Spacecraft atomic-line measurements \\
P7 & Photon group velocity decrease at high $\omega$ & GRB-time-lag scaling as $\omega^2$ & Constrains $a \lesssim 10^{-19}\,\text{m}$ \\
P8 & Atomic spectrum gap inside Roche radius & gradient $g_\text{crit}$ for hydrogen & Neutron-star photosphere spectroscopy \\
P9 & Concordance of optical axis $(1,1,-1)$ across kinematic, gauge, gravitational, and atomic channels & inter-channel direction agreement at $(a/\lambda)^2$ scale & GRB / CMB / LIGO / atomic-clock direction-resolved binning \\
\hline
\end{tabular}
\end{center}

\paragraph{The calibration window.}
P7 already constrains the lattice spacing from above:
$a \lesssim 10^{-19}\,\text{m}$. The Compton-calibration numbers in
the table above are therefore upper bounds. The framework's
surviving calibration window is $\ell_P \lesssim a \lesssim 10^{-19}\,\text{m}$,
a 16-decade range. The numerical signal strengths of P1–P6 within
this window are smaller than the Compton-scale entries by factors
$(a/\lambda_C)^n$ where $n \in \{1, 2\}$ depending on the prediction.
The qualitative signatures (a discrete time-dilation cutoff in P2,
a discontinuous Stark threshold in P5, a hexapole CMB pattern in
P3, etc.) survive at any calibration; only their amplitudes scale
with $a$.

This is the most useful kind of prediction: an existing measurement
(GRB time delays) has already bounded the framework's parameter
space. Future improvements in P5 (Stark-threshold spectroscopy),
P2 (optical-clock comparisons), and P3 (CMB anisotropy searches)
will narrow the window further. The framework is falsifiable in
detail.

\paragraph{What would falsify the framework as a whole.}
The framework is not a one-shot prediction; it is a structural
account of quantum mechanics, gravity, and electromagnetism on a
common lattice substrate. To falsify the framework as a whole would
require either:

\begin{itemize}
    \item Demonstrating that one of the derivations of earlier
    sections (the Dirac equation in §6, Newton's gravity in §7,
    Maxwell's equations in §9) does not hold. None of these has
    been challenged to date.
    \item Confirming \emph{all} nine predictions to better than
    the predicted precision, with no positive deviation from
    standard physics. This would force the lattice spacing to be
    smaller than the Planck length, at which point the framework's
    correspondence with established physics is preserved but its
    discrete substrate is unobservable in principle.
\end{itemize}

The first kind of falsification would mean the framework is wrong
in its derivations; the second would mean the framework is right
but undetectable. Either is a meaningful outcome.
