% ============================================================
% vacuum_twist_field_equations.tex
% Section 9: The Vacuum Twist: Gravity, Electromagnetism, and the
%            Unified Phase Field
% ============================================================

\label{sec:vacuum_twist}

Standard physics treats gravity and electromagnetism as independent
forces. They have different field contents (a metric tensor vs.\ a
vector potential), different gauge structures (diffeomorphisms vs.\
$U(1)$), and different reasons to exist (general covariance vs.\
phase invariance of charged matter). That they are both consistent at
the same point in spacetime is a coincidence the standard model leaves
unaddressed.

The $\unity$ framework dissolves the coincidence.
On the lattice, the two long-range forces are not independent --- they
are the two channels of a Helmholtz decomposition of the phase
gradient field, and both are populated by the same bipartite tick
rule. The divergence channel produces gravity
(derived end-to-end in Section~\ref{sec:gravity}); the curl channel
produces electromagnetism (derived in this section, with the
matter-side coupling forced by the tick rule and the field-side
dynamics following from the standard lattice-gauge action of Wilson).

The unification is not at the level of a single field equation. It is
at the level of \emph{substrate and coupling}: both forces live on the
same lattice, both couple to matter through the same Zitterbewegung
parameter $\omega$, and both reduce to known physics in their
continuum limits. What differs is which channel of the Helmholtz
decomposition each populates.

% ── 9.1 The Two Channels: Helmholtz Decomposition ────────────────────────────
\subsection{The Two Channels of Vacuum Deformation}
\label{subsec:two_channels}

A theorem of Helmholtz states that every sufficiently smooth vector
field $\mathbf{v}(\mathbf{x})$ on $\mathbb{R}^3$ decomposes uniquely
into a gradient part and a curl part:
\begin{equation}
    \mathbf{v} = -\nabla\Phi + \nabla\times\mathbf{A}.
    \label{eq:helmholtz}
\end{equation}
The gradient part has $\nabla\times\mathbf{v} = 0$; the curl part has
$\nabla\cdot\mathbf{v} = 0$. These are \emph{the} two long-range
degrees of freedom of any smooth field on flat space, because the
Laplacian factors as
$\nabla^2 = \nabla(\nabla\cdot\,) - \nabla\times(\nabla\times\,)$.
There is no third channel.

Applied to the phase gradient field on the lattice, the two channels
have distinct physical character.
The divergence channel
$\nabla\cdot(\nabla\phi) = \nabla^2\phi$ measures sources and sinks of
phase; it is a scalar and (near a localised source) always carries one
sign. The curl channel records circulation of phase around closed
loops; it is a tensor and admits two chiralities, corresponding to
the two sublattices of the bipartite $\Tdiamond$ structure.

The lattice populates both channels through the same tick rule. The
divergence channel is populated by clock-density gradients --- a
region with more registered sessions has more bookkeeping per
macro-tick, and the resulting bias in the hop probability produces a
scalar potential whose Laplacian is the source. The curl channel is
populated by phase circulation around plaquettes --- a region where
the relative phase between sublattices winds non-trivially around a
closed lattice loop carries a tensor field strength whose source is
the chiral matter current.

The two channels do not interfere with each other. Gravity acts on
all matter (every session has a clock that contributes to
$\rho_\text{clock}$); electromagnetism acts only on chiral matter
(only sessions with a non-zero RGB/CMY imbalance couple to $A_\mu$).
This selectivity is geometric, not phenomenological.

% ── 9.2 The Divergence Channel: Gravity ──────────────────────────────────────
\subsection{The Divergence Channel: Gravity (Recap)}
\label{subsec:gravity_recap}

The divergence-channel derivation is given end-to-end in
Section~\ref{sec:gravity}. We summarise it here only for parallelism.

The hop probability $p_\text{hop} = \cos^2(\delta\phi/2)$ with
$\delta\phi = \omega + V$ expands in the weak-field limit as
$p_\text{hop} \approx p_0 - m\,V$, where $V \propto -\rho_\text{clock}$
is the topological potential induced by the clock-density gradient.
Equilibrium of the resulting probability flux gives the Boltzmann
distribution
\begin{equation}
    \rho_\text{clock}(\mathbf{x}) = \bar\rho\,
    \exp\!\bigl(-\phi_\text{Newton}(\mathbf{x})/c^2\bigr),
    \label{eq:boltzmann_clock_recap}
\end{equation}
which linearises to Poisson's equation
$\nabla^2\phi_\text{Newton} = 4\pi G\,\rho_\text{mass}$ in the weak
field. Newton's law of gravitation
$\mathbf{F} = -GMm\,\hat{\mathbf{r}}/r^2$ follows from the Green's
function of the Laplacian.

Every step uses only the existing tick rule. No additional field is
introduced. Gravity is, in this framework, the effective dynamics of
clock density on the multi-session scheduler.

% ── 9.3 The Curl Channel: Matter-Side Derivation ─────────────────────────────
\subsection{The Curl Channel: Matter-Side Derivation}
\label{subsec:em_matter_side}

The curl channel requires more architecture than the divergence
channel, because the natural field carrying the curl is a vector
$A_\mu$ rather than a scalar.
This subsection derives the matter-side structure: the existence of
the gauge field, its transformation law, the field strength $F_{\mu\nu}$,
and the matter current $J^\mu$. Each is forced by the existing tick
rule with no additional postulates.

\paragraph{The link variable is forced.}
Each causal session carries a $U(1)$ phase
(Section~\ref{subsec:phase_oscillator}). When a session hops from
$\mathbf{x}$ to $\mathbf{x} + a\hat{\mu}$, its phase acquires a factor
$\exp(i\,a\,A_\mu(\mathbf{x}))$ in addition to the Zitterbewegung
phase $\delta\phi$. This is the Peierls substitution, already present
in the lattice tick rule (\texttt{CausalSession.\_kinetic\_hop}).
The natural object encoding this phase is the link variable
\begin{equation}
    U_\mu(\mathbf{x}) = e^{i\,a\,A_\mu(\mathbf{x})} \in U(1),
    \label{eq:link_variable}
\end{equation}
living on the link from $\mathbf{x}$ to $\mathbf{x} + a\hat{\mu}$.
$U_\mu$ is the most general $U(1)$-valued local object that can act
on a hop without breaking locality. Its existence and its $U(1)$
group structure are forced by the matter content (every session has a
$U(1)$ phase, so the relative phase between sessions on adjacent
nodes is a $U(1)$ element); they are not additional postulates.

\paragraph{Gauge covariance of the tick rule.}
Under a local $U(1)$ gauge transformation
$\Lambda(\mathbf{x})$, the session phase, the gauge field, and the
link variable transform as
\begin{align}
    \psi(\mathbf{x}) &\to e^{i\Lambda(\mathbf{x})}\,\psi(\mathbf{x}),
        \label{eq:psi_gauge} \\
    A_\mu(\mathbf{x}) &\to A_\mu(\mathbf{x}) + \partial_\mu\Lambda(\mathbf{x}),
        \label{eq:A_gauge} \\
    U_\mu(\mathbf{x}) &\to e^{i\Lambda(\mathbf{x}+a\hat{\mu})}\,U_\mu(\mathbf{x})\,e^{-i\Lambda(\mathbf{x})}.
        \label{eq:U_gauge}
\end{align}
The transformation of $U_\mu$ follows from
$U_\mu = e^{i\,a\,A_\mu}$ and
$a\,\partial_\mu\Lambda \approx \Lambda(\mathbf{x}+a\hat{\mu}) - \Lambda(\mathbf{x})$
in the small-$a$ limit. Applying the transformation to the
Peierls-coupled tick rule
$\psi(\mathbf{x}+a\hat{\mu}) = U_\mu(\mathbf{x})\cdot(\text{tick}\,\psi)(\mathbf{x})$
shows that the rule is gauge covariant: the new $\psi$ at
$\mathbf{x}+a\hat{\mu}$ acquires the local phase
$e^{i\Lambda(\mathbf{x}+a\hat{\mu})}$ exactly. Physical observables
($|\psi|^2$, phase differences, $\unity$) are gauge invariant by
construction.

\paragraph{The plaquette is the field strength.}
The smallest non-trivial closed loop on the lattice is a plaquette in
the $(\mu, \nu)$ plane, traversed
$\mathbf{x} \to \mathbf{x}+a\hat{\mu} \to \mathbf{x}+a\hat{\mu}+a\hat{\nu} \to \mathbf{x}+a\hat{\nu} \to \mathbf{x}$.
The $U(1)$ holonomy around this loop is the Wilson loop
\begin{equation}
    W_{\mu\nu}(\mathbf{x}) =
        U_\mu(\mathbf{x})\,U_\nu(\mathbf{x}+a\hat{\mu})\,
        U_\mu^*(\mathbf{x}+a\hat{\nu})\,U_\nu^*(\mathbf{x}).
    \label{eq:wilson_loop}
\end{equation}
Expanding $A_\mu(\mathbf{x}+a\hat{\nu}) \approx A_\mu(\mathbf{x}) + a\,\partial_\nu A_\mu$ for small $a$:
\begin{equation}
    W_{\mu\nu}(\mathbf{x})
        \xrightarrow{a\to 0}
        \exp\!\bigl(i\,a^2\,F_{\mu\nu}(\mathbf{x})\bigr),
    \label{eq:wilson_to_F}
\end{equation}
with
$F_{\mu\nu} = \partial_\mu A_\nu - \partial_\nu A_\mu$ the standard
field strength tensor.
$W_{\mu\nu}$ is gauge invariant: the corner gauge transformations of
Eq.~\eqref{eq:U_gauge} cancel pairwise around the loop. So
$F_{\mu\nu}$ emerges in the continuum limit as the smallest
non-trivial gauge-invariant local operator built from link variables
--- without postulating it.

\paragraph{The Dirac current.}
The bipartite tick rule with Peierls links is equivalent in the
continuum limit to the lattice Dirac operator
\begin{equation}
    D_\text{lat}\,\psi(\mathbf{x}) = \frac{1}{2a}\sum_\mu \gamma^\mu
    \bigl[
        U_\mu(\mathbf{x})\,\psi(\mathbf{x}+a\hat{\mu})
        - U_\mu^*(\mathbf{x}-a\hat{\mu})\,\psi(\mathbf{x}-a\hat{\mu})
    \bigr],
    \label{eq:lattice_dirac}
\end{equation}
where the $\gamma^\mu$ matrices are the ones derived in
Section~\ref{subsec:dirac_limit} from the bipartite RGB/CMY
structure (the equivalence between the alternating tick rule and the
compact Dirac operator is the standard staggered-fermion result of
Kogut and Susskind~\cite{kogutsusskind1975}).
The matter action $S_\text{matter} = \sum_\mathbf{x} a^4\,\bar\psi(\mathbf{x})\,D_\text{lat}\,\psi(\mathbf{x})$
varied with respect to $A_\mu$, using $\delta U_\mu = i\,a\,U_\mu\,\delta A_\mu$, gives
\begin{equation}
    \frac{\delta S_\text{matter}}{\delta A_\mu(\mathbf{x})}
        = a^4\,\bar\psi(\mathbf{x})\,\gamma^\mu\,\psi(\mathbf{x}) + O(a),
\end{equation}
which in the continuum limit is the Dirac current
\begin{equation}
    \boxed{\;J^\mu(\mathbf{x}) = \bar\psi(\mathbf{x})\,\gamma^\mu\,\psi(\mathbf{x}).\;}
    \label{eq:dirac_current}
\end{equation}
The matter current is forced by the bipartite structure; it is not
postulated.

% ── 9.4 Maxwell's Equations from the Wilson Action ───────────────────────────
\subsection{Maxwell's Equations from the Wilson Action}
\label{subsec:maxwell_from_wilson}

The matter-side derivation of Section~\ref{subsec:em_matter_side}
gives us the link variable, the field strength, and the matter
current. To obtain a closed dynamical system we need an equation of
motion for $A_\mu$ itself, which requires an action for the gauge
field.

The framework as currently formulated does not derive this action
from the tick rule directly. Section~\ref{subsec:open_program} below
sets out a programme by which the gauge action could be \emph{induced}
from matter histories rather than postulated; that programme is
left to a follow-on paper. For the present section we adopt the
standard result of lattice gauge theory: locality, gauge invariance,
minimality, and reflection positivity uniquely fix the gauge action
to the Wilson form~\cite{wilson1974}
\begin{equation}
    S_\text{gauge}[U] = \frac{1}{g^2}
        \sum_\text{plaquettes}\bigl(1 - \mathrm{Re}\,W_{\mu\nu}(\mathbf{x})\bigr).
    \label{eq:wilson_action}
\end{equation}
Using the small-$a$ expansion
$1 - \mathrm{Re}\,W_{\mu\nu} \approx 1 - \cos(a^2 F_{\mu\nu})
\approx \tfrac{1}{2}a^4\,F_{\mu\nu}F^{\mu\nu}$, the continuum limit
of Eq.~\eqref{eq:wilson_action} is
\begin{equation}
    S_\text{gauge} \;\xrightarrow{a\to 0}\;
        \frac{1}{2g^2}\int d^4x\,F_{\mu\nu}\,F^{\mu\nu},
    \label{eq:maxwell_action}
\end{equation}
the standard Maxwell action (with $g^2 = 2$ in our normalisation).
Varying $S_\text{gauge} + S_\text{matter}$ with respect to $A_\mu$
gives
\begin{equation}
    \boxed{\;\partial_\nu F^{\nu\mu} = J^\mu,\;}
    \label{eq:maxwell_inhomogeneous}
\end{equation}
the inhomogeneous Maxwell equation. The homogeneous equation
$\partial_{[\rho}F_{\mu\nu]} = 0$ is automatic from
$F_{\mu\nu} = \partial_\mu A_\nu - \partial_\nu A_\mu$.

% ── 9.5 Reductions and Numerical Confirmation ────────────────────────────────
\subsection{Reductions and Numerical Confirmation}
\label{subsec:em_reductions}

Maxwell's equations as derived above reduce to the standard textbook
results in their respective limits.

\paragraph{Massless photon (vacuum).}
With $J^\mu = 0$, Eq.~\eqref{eq:maxwell_inhomogeneous} becomes
$\Box A^\mu - \partial^\mu(\partial\cdot A) = 0$.
In Lorenz gauge $\partial\cdot A = 0$ this is the photon wave
equation $\Box A^\mu = 0$. The photon is transverse with two
polarisation states, corresponding to the two RGB/CMY chiralities of
the lattice. Spin-$1$, masslessness, and gauge invariance are
structural consequences of the derivation; none is postulated
separately.

\paragraph{Static charge (Coulomb's law).}
With $\partial_t = 0$ and $\mathbf{J} = 0$, the spatial part of
Eq.~\eqref{eq:maxwell_inhomogeneous} reduces to
$\nabla\cdot\mathbf{E} = \rho$. The Green's function of the Laplacian
gives
$\mathbf{E} = q\,\hat{\mathbf{r}}/(4\pi r^2)$ for a point charge,
which is Coulomb's law.

\paragraph{Static current (Biot--Savart).}
With $\partial_t = 0$ and $\rho = 0$, the spatial part is
$\nabla\times\mathbf{B} = \mathbf{J}$. The Green's function gives the
Biot--Savart law for the magnetic field of a current distribution.

\paragraph{Numerical confirmation (\texttt{exp\_08}, \texttt{exp\_09},
\texttt{exp\_19c}).}
The qualitative structure of the curl channel is confirmed by
\texttt{exp\_08}: a charged session in a dipole-aligned background
deflects \emph{tangentially} (curl-like) at $33^\circ$ from the
applied field and $30^\circ$ alignment with $\mathbf{V}\times\hat{z}$,
while a session in a clock-density gradient deflects \emph{radially}
(divergence-like, $-3.55$ nodes inward).
The two channels are visibly distinct and behave as the geometry
predicts.
The photon dispersion relation derived from the Wilson action matches
the lattice harmonic spectrum measured in \texttt{exp\_09}: linear
band crossing at $k\to 0$ with the lattice cutoff modifying the
high-frequency tail. Photon emission in the bipartite system,
demonstrated in \texttt{exp\_19c} and discussed in
Section~\ref{subsec:emission}, is the dynamical creation of a third
session that closes the $\unity$ accounting when the
electron--proton joint resonance jumps Arnold tongues. All three
experiments confirm the curl-channel geometry.

% ── 9.6 Honest Scope and the Path Forward ────────────────────────────────────
\subsection{Honest Scope: What is Derived and What is Postulated}
\label{subsec:honest_scope}

The derivation of this section is asymmetric with the gravity
derivation of Section~\ref{sec:gravity}, and it is worth being
explicit about the asymmetry.

The gravity derivation introduces no new field beyond what the tick
rule already provides. Clock density $\rho_\text{clock}$ is a
property of the multi-session scheduler; the divergence-channel
dynamics is derived end-to-end from the existing tick rule.

The electromagnetism derivation introduces the link variable
$A_\mu$ as a $U(1)$-valued field on lattice links. Its existence and
$U(1)$ group structure are forced by the matter content; its
coupling to matter through Peierls substitution is forced by the
existing tick rule; the field strength $F_{\mu\nu}$ and the Dirac
current $J^\mu$ are forced by gauge invariance and the bipartite
structure respectively. So far, nothing is postulated beyond the
existing tick rule.

What \emph{is} postulated is the Wilson action of
Eq.~\eqref{eq:wilson_action}, the standard lattice-gauge prescription
that closes the dynamical system. The Wilson action is uniquely fixed
by locality, gauge invariance, minimality, and reflection positivity,
but those criteria are themselves additional postulates external to
the tick rule.

In summary:

\begin{center}
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{lll}
\hline
\textbf{Channel} & \textbf{Substrate} & \textbf{Field equation} \\
\hline
Divergence (gravity) & clock-density gradient & derived from tick rule \\
Curl (electromagnetism) & $U(1)$ link variable & structural form derived \\
                       &                       & (Appendix~\ref{sec:induced_gauge_action}); \\
                       &                       & numerical prefactor open \\
\hline
\end{tabular}
\end{center}

The unification claim of this section survives the asymmetry, because
the unification is at the level of substrate and coupling rather than
at the level of a single field equation:

\begin{itemize}
    \item Both forces share the same matter substrate (the bipartite
    Zitterbewegung structure with $\unity$).
    \item Both forces couple to matter through the same mechanism
    (the topological-potential modification of the hop probability,
    which manifests as a clock-density gradient for gravity and as a
    Peierls phase for electromagnetism).
    \item Both forces reduce to known physics in their continuum
    limits (Newton from divergence, Maxwell from curl).
\end{itemize}

This is more than standard physics offers --- there, gravity and EM
are independent theories with no shared substrate.

% ── 9.7 Open: induced gauge action ───────────────────────────────────────────
\subsection{Open Program: The Induced Gauge Action}
\label{subsec:open_program}

The natural route to closing the asymmetry of
Section~\ref{subsec:honest_scope} is the Sakharov--Zeldovich induced
action programme~\cite{sakharov1967}: the gauge action of
Eq.~\eqref{eq:wilson_action} should not be postulated but should be
\emph{induced} by integrating out matter sessions in a path-integral
formulation of the tick rule.

Schematically, with $Z[U]$ the matter partition function in
background $A_\mu$ and $S_\text{eff}[U] = -\log Z[U]$ the induced
effective action,
\begin{equation}
    Z[U] = \det\bigl(D_\text{lat}[U]\bigr),
    \qquad
    S_\text{eff}[U] = -\mathrm{Tr}\,\ln D_\text{lat}[U].
    \label{eq:induced_action}
\end{equation}
The standard expansion of $\mathrm{Tr}\,\ln D_\text{lat}[U]$ as a sum
over closed lattice loops, combined with the locality + gauge
invariance + minimality arguments, dominates the effective action at
leading order with the Wilson plaquette form. This is a calculation
that has been carried out in standard lattice gauge
theory~\cite{kogutsusskind1975}; in the present framework it has
not yet been done.

If the induced-action calculation succeeds, it converts the Wilson
action from a postulate into a derived consequence of the bipartite
tick rule under matter integration. The unification claim of this
section then becomes single-equation: gravity from the divergence
channel of the same tick rule that, through matter integration,
generates the curl-channel dynamics.

If the calculation produces a coupling of unexpected sign or magnitude
--- or higher-loop terms that do not match standard QED at the
expected order --- that is a falsifiable result. The induced
$\beta$-function should match the QED $\beta$-function in the
appropriate limit; deviations would constitute a prediction.

Three concrete tasks remain to close the gap:

\begin{enumerate}
    \item Construct a path-integral measure from the stochastic tick
    rule. The framework already evolves complex spinors unitarily, so
    the path integral exists implicitly --- this is a writing-down
    exercise rather than a research-level question.
    \item Compute the matter determinant
    $\mathrm{Tr}\,\ln D_\text{lat}[U]$ explicitly, with the
    bipartite RGB/CMY structure giving the specific $\gamma^\mu$
    realisation. Standard lattice gauge theory machinery applies.
    \item Extract the classical limit via stationary phase in the
    small-$a$ expansion. Standard WKB in the lattice spacing.
\end{enumerate}

Of these three tasks, the \emph{structural} step -- showing that
the leading-order term of $\mathrm{Tr}\,\ln D_\text{lat}[U]$, when
expanded in closed lattice loops, has exactly the bipartite plaquette
form -- is carried out symbolically in
Appendix~\ref{sec:induced_gauge_action}.  The result is that the
leading induced action takes the form
\begin{equation*}
    S_\text{eff}^{(\text{leading})}[A]
        = \frac{c\,a^4}{2}\,\sum_{a<b}\bigl(V_a^i V_b^j F_{ij}\bigr)^2,
\end{equation*}
with a tensor coefficient inheriting the bipartite anisotropy
(eigenvalues $\{4, 4, 16\}$, optical axis $(1,1,-1)$) and recovering
the standard Maxwell density on $O_h$-averaged observables.  The
operator-level residual is a closed-form gauge-sector birefringence
with the same optical axis as the kinematic-sector prediction P7
(Section~\ref{subsec:p7_photon_dispersion}).
What remains is the explicit computation of the prefactor $c = 1/g^2$,
which requires the full one-loop $\mathrm{Tr}\,\ln$ calculation
including standard staggered-fermion doubler bookkeeping; that is the
paper-length follow-on project, with its own verification programme
against QED precision tests ($g{-}2$, the Lamb shift, the
$\beta$-function running).

The current paper therefore claims the matter-side derivation
honestly (Sections~\ref{subsec:em_matter_side}--\ref{subsec:em_reductions})
and treats the Wilson action as the standard lattice-gauge postulate
(Section~\ref{subsec:maxwell_from_wilson}).  The induced-action
programme is closed at the structural level by
Appendix~\ref{sec:induced_gauge_action} and reduced to a single
remaining computation (the numerical prefactor) that does not affect
the unification claim of this section.
