\documentclass[12pt, letterpaper]{article}

%% Load packages -- build from the paper/ directory:
%%   cd paper && pdflatex main.tex
%% or set TEXINPUTS if building from repo root.
\input{macros/packages}
\input{macros/commands}

%% Page geometry (also set here as fallback in case packages.tex path fails)
\geometry{top=1in, bottom=1in, left=1in, right=1in}

\title{\textbf{Beta -- Version 0.98-RC}\\[0.5em]
       Geometry First:\\
       Quantum Mechanics, Gravity, and the Origin of the Standard Model\\
       from a Single Conservation Law}
\author{Jack Menendez\\
        \small Eugene, OR\\
        \small ORCID: \href{https://orcid.org/0009-0003-1166-307X}{0009-0003-1166-307X}\\
        \small \href{mailto:jackdmenendez@gmail.com}{jackdmenendez@gmail.com}}
\date{\today}

\begin{document}

\maketitle

\noindent
\href{https://doi.org/10.5281/zenodo.20017133}%
     {\texttt{Zenodo DOI: 10.5281/zenodo.20017133}}\quad
\href{https://github.com/JackDMenendez/discrete-causal-lattice}{GitHub: JackDMenendez/discrete-causal-lattice}

\noindent\textbf{Disclaimer:} This is a working paper, and the content herein is subject to change. The author welcomes feedback and constructive criticism. Please reach out via email or social media with any comments or questions.

\medskip
\noindent\textbf{License.}  Paper text and figures: Creative Commons
Attribution 4.0 (CC BY 4.0).  Source code in the companion
repository: MIT License.

\medskip
\noindent\textbf{Data availability.}  All numerical experiments,
data files, and the LaTeX source for this paper are publicly
available at
\href{https://github.com/JackDMenendez/discrete-causal-lattice}%
     {\texttt{github.com/JackDMenendez/discrete-causal-lattice}}.
The exact commit corresponding to this release is the v0.98-RC tag,
archived at
\href{https://doi.org/10.5281/zenodo.20017133}%
     {\texttt{doi:10.5281/zenodo.20017133}}.
A compact reproducibility quickstart, including a 10-line minimal
verification snippet, is in
Appendix~\ref{sec:reproducibility}.

\medskip
\noindent\textbf{Keywords.}  discrete spacetime; bipartite causal
lattice; single conservation law; emergent quantum mechanics;
clock-fluid gravity; Arnold-tongue quantization; induced gauge
action; lattice birefringence.

\bigskip
\noindent\textbf{What changed in v0.98-RC (since v0.97-RC).}
v0.98-RC is an experimental-progress and content-alignment release.
Two new experiments completed their first runs; two new entries were
added to the audit table (Table~\ref{tab:audit}, both \texttt{PART});
and the paper text was systematically reconciled with the audit-table
statuses, with the help of an automated consistency-check pass
introduced in this release:

\begin{itemize}
\item \textbf{\texttt{exp\_20} --- three-arm emission-operator
    comparison.}  First experiment to test whether a unitary 2-mode
    rotation (beam splitter, row 6 of the operation algebra) is the
    correct emission operator for the lattice's joint $\mathcal{A}=1$
    conservation, compared directly against the as-written drain in
    \texttt{exp\_19c} (arm A) and the phase-rotation drain (arm C).
    Joint $\mathcal{A}=1$ preserved to machine precision
    ($8.9\!\times\!10^{-16}$) under the beam splitter; the
    \texttt{exp\_19c} drain reproduces the row-5 non-unitarity
    arithmetic exactly.  Phase-dependent non-monotonic transfer in
    arm B is an unanticipated refinement to the operation-algebra
    table that a follow-on calculation will need to address.
    Orbital settling not achieved at the swept rate; see next item.
\item \textbf{\texttt{exp\_12b} --- long-duration two-body baseline
    (no emission).}  The bare \texttt{exp\_12} setup, run for 6000 ticks.
    $r_\text{peak}$ escapes from $R_1$ to grid edge ($\sim 83$) by
    tick $\sim 2000$ and never returns.  $\mathcal{A}=1$ preserved at
    machine precision throughout.  The orbital escape seen across all
    three arms of \texttt{exp\_20} is inherited from this baseline,
    not caused by the emission operator.  Refines but does not
    invalidate \texttt{exp\_12}'s 4-sig-fig PASS, which is scored on
    a shorter $k$-scan resonance peak.
\item \textbf{Two new entries in the audit table} (both
    \texttt{PART}, both additive --- no existing entries downgraded):
    ``Two-body long-horizon stability'' (the \texttt{exp\_12b}
    finding) and ``Emission-operator joint $\mathcal{A}=1$
    conservation'' (the \texttt{exp\_20} arm B finding).  The
    evidence column for the existing entry ``Photon emission as
    $\mathcal{A}=1$ necessity'' was updated to reflect the
    \texttt{exp\_20} results; that entry remains \texttt{PART}.
\item \textbf{Paper-text reconciliation.}  Four sites with
    pre-\texttt{exp\_20} \texttt{exp\_19c} overstatement aligned with
    the audit-table: \texttt{abstract.tex} items (v) and (vi),
    \texttt{introduction.tex} ($\times 2$ mentions),
    \texttt{hydrogen\_spectrum.tex}, and
    \texttt{vacuum\_twist\_field\_equations.tex}.  The row-5
    amplitude-drain non-unitarity arithmetic
    $(1 - 2m + 2m^2)\lVert\psi_e\rVert^2$ is now explicitly cited in
    the hydrogen-spectrum write-up of \texttt{exp\_19c}.
\item \textbf{New introduction paragraph: ``One law at three
    resolutions.''}  Identifies the clock fluid density
    $\rho_\phi(x) = \sum_i \omega_i\,|\psi_i(x)|^2$ as the
    $\omega$-weighted form of the joint amplitude density
    $\mathcal{A}_\text{joint}(x) = \sum_i |\psi_i(x)|^2$.  Probability
    conservation, mass-energy conservation, and gravity all share one
    structural origin --- sharpens the framework's ``single
    conservation law'' claim from assertion to identification.
\end{itemize}

An automated consistency-check tool
(committed at \texttt{.claude/agents/dcl-claim-auditor.md}) was added
to compare paper prose against the audit-table statuses; during this
release cycle it identified two further overstatements in the abstract
that the manual sweep had missed.  Eight follow-on or speculative
notes were moved out of the project repository into a separate
private archive, leaving \texttt{notes/} at 53 files (down from 61),
each tied to a main-paper claim or an active experiment.

A more detailed change log lives in
\texttt{release\_notes/v0.98-RC.md} in the project repository.
Earlier versions remain available at the Zenodo deposit's prior
DOIs.

\begin{abstract}
    \input{sections/abstract}
\end{abstract}

\newpage
\tableofcontents
\newpage

%% ============================================================
%% PART I: FOUNDATIONS
%% ============================================================
\section{Introduction}
\input{sections/introduction}

\section{The Octahedral Substrate \texorpdfstring{($\Tdiamond$)}{(T\_diamond\textasciicircum 3)}}
\input{sections/octahedral_substrate}

\section{Causal Sessions and the Phase Oscillator}
\input{sections/causal_sessions}

%% ============================================================
%% PART II: DYNAMICS
%% ============================================================
\section{The Tick Scheduler}
\input{sections/tick_scheduler}

\section{Phase Propagation and the Unity Constraint}
\input{sections/phase_propagation}

\section{Emergent Kinematics on the \texorpdfstring{$\Tdiamond$}{T\_diamond\textasciicircum 3} Lattice}
\input{sections/emergent_kinematics}

\section{Gravity as Clock Density}
\input{sections/gravity_as_clock_density}

\section{Clock Fluid Dynamics and the Continuum Limit}
\input{sections/clock_fluid_dynamics}

\section{The Vacuum Twist: Gravity, Electromagnetism, and the Unified Phase Field}
\input{sections/vacuum_twist_field_equations}

%% ============================================================
%% PART III: QUANTUM PHENOMENA
%% ============================================================
\section{Interference and the Huygens Lantern}
\input{sections/interference}

\section{The Observer as a Clock}
\input{sections/observer_as_clock}

%% ============================================================
%% PART IV: PREDICTIONS AND VERIFICATION
%% ============================================================
\section{Path Counting and Falsifiable Predictions}
\input{sections/predictions}

\section{Lattice Harmonics and Quantization}
\input{sections/lattice_harmonics}

\input{sections/fractal_universe}

\section{The Hydrogen Spectrum from Lattice Geometry}
\input{sections/hydrogen_spectrum}

\section{Conclusion}
\input{sections/conclusion}

%% ============================================================
%% APPENDICES
%% ============================================================
\appendix

\section{Code and Data}
\input{sections/code_and_data}

\section{The Induced Gauge Action}
\input{sections/induced_gauge_action}

\section{Reproducibility Quickstart}
\input{sections/reproducibility}

%% ============================================================
\bibliographystyle{unsrt}
\bibliography{paper-bib/references}

\end{document}
